### 2022-02-23 manuscript/rebuttal_letter.Rmd
Initial commit
Post-revision analysis pipeline

---
title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
subtitle: "Point-to-Point Reply"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  pdf_document:
    latex_engine: xelatex
bibliography: cov_biblio.bib
csl: european-journal-of-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H}\floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = 'cairo_pdf')

```

\newpage

# Reviewer 1

## Comments to the Author

This study by Haschka et al. entitled “Early alterations of blood monocyte subsets and transferrin receptor expression are linked to infection severity in hospitalized COVID-19 patient” is one of the numerous studies that have phenotyped the blood samples from COVID-19 patients. This study applies a similar approach as described in previous studies, i.e. investigating differences in surface marker expression in myeloid cells aiming to find the prognostic markers for early diagnosis of severe disease. There are several concerns that require addressing to strengthen the manuscript further.

## Reply to the Reviewer

We thank the Reviewer for careful study of the manuscript, appreciation of our work and valuable feedback. Please find below the replies to the particular issues.

## Issue 1

The title “Early alterations of blood monocyte subsets and transferrin receptor expression are linked to infection severity in hospitalized COVID-19 patient” could be misleading. (1) How do the authors establish as an early stage? Are the writers aware of what "late" modifications imply and how they differ from "early" alterations? Can the authors elaborate on what "late" modifications entail and differ from "early" alterations?

## Reply 1

David, könntest Du was draften? Sollen wir irgendwie erwähnen, dass wir auch weitere Zeitpunkte bei manchen Patienten gemessen haben? Oder den Titel und die Formulierung 'early' ändern?

## Issue 2

Transferrin receptor expression (CD71) showed a low odds ratio of 0.92 (Figure 6) while there are more significant findings such as int monocyte CD64 expression. Why did the authors choose transferrin receptor expression? Can the authors elaborate more on this?

## Reply 2

In the initial version of the manuscript, the choice of the variables for the multi-parameter score was done with LASSO algorithm (least absolute shrinkage and selection operator) [@Tibshirani1996]. After discussions in the study team, involving a data scientist (Piotr Tymoszuk), we decided to abstain from predictive modeling in the revised manuscript. The primary reason for that was the low number of COVID-19 participants (n = `r mono_data$data_tbl %>% count(covid_status) %>% .$n %>% .[2]`) making the risk multi-parameter risk modeling highly ambiguous and sensitive to any changes in the gating strategy.

In general, to make the analysis more comprehensive and reproducible, we introduced several changes. 
First, we restrict, the analysis cohort to the uniform subset of the initially enrolled patients for which high quality flow cytometry data could be obtained (__Figure 1__).
Second, Kruskal-Wallis test was used to identify cytometry features of interest differentially regulated between healthy controls, moderate and severe COVIV-19 patients (__Figure 2__). 
Third, the p values were consistently corrected for multiple testing with the Benjamini-Hochberg method. 
Fourth, correlation analysis of single cytometry features and markers of systemic inflammation were replaced by a more informative clustering analysis. 
Finally, instead of predictive modeling, we focused on a potentially more interesting and reproducible interplay between myeloid immunity phenotype, inflammation and COVID-19 severity. To this end, we subjected the study participants to a clustering analysis in respect to all recorded flow cytometry variables (__Table S3__) using the self-organizing map (SOM) and hierarchical clustering technique (__Figure S9__) [@Kohonen1995; @Wehrens2018; @Vesanto2000]. Of note, such neuronal network-based technique is well suited for analysis of multi-dimensional data sets with multiple inter-correlated variables and was applied by in a similar setting in a recent paper from our division [@Sonnweber2022]. By this means, we could identify four distinct clusters of the study participants (__Figure 4__, __Figure S9C__):

* Cluster #1 consisting of predominantly (75%) healthy controls and moderate COVID-19 patients without pathological neutrophil expansion, preserved monocyte subset distribution as well as 'healthy-like' levels of myeloid CD64 and CD86. COVID-19 individuals assigned to this cluster had the lowest levels of serum inflammation markers IL6, CRP and neopterin (__Figure 5__).

* Cluster #2 characterized by an upregulation of myeloid leukocyte CD86, monocyte CD163, expansion of non-classical monocytes. This cluster included both moderate and severe COVID-19 cases and substantially higher levels of neopterin and CRP than cluster #1 (__Figure 5__).

* Cluster #3 included predominantly (73%) severe COVID-19 patients, displayed an upregulation of myeloid leukocyte CD64, monocyte CD40, CD71 and CD274 (PD-L1) and a depletion of non-classical monocytes along with high levels of IL6, CRP and neopterin (__Figure 5__).

* Cluster #4 consisted almost exclusively of severe COVID-19 cases and was characterized by high levels of systemic inflammation, a strong expansion of neutrophils and an upregulation of neutrophil CD64.

In addition, the above clusters tended to differ in iron turnover parameters. In particular, in the clusters #2, #3 and #4, low values of plasma iron and transferrin saturation were observed, suggestive of inflammatory iron restriction. Importantly, we could not find any evident differences in base demographic variables: age, sex distribution and body mass index between the clusters (__Table S4__). Cumulatively, we interpret these clusters as separate patterns of innate response to the pathogen and systemic inflammation which may be independent of the demographic background of the COVID-19 patient. In the discussion, we elaborate on the relation between the clusters and the patterns of adequate immune response (Cluster #1), immunoregulation or immunosuppression (Cluster #3), hyper-activation phenotype (Cluster #3) and emergency myelopoiesis (Cluster #4) described in the literature [@Silvin2020; @Schulte-Schrepping2020; @Christensen2021; @Kvedaraite2021].

## Issue 3

CD64 expression on neutrophils is a common known marker for detecting sepsis in patients (Ng et al., 2004; Sack, 2017). It can be up-regulated similarly with CD14 and TLRs. The authors should mention this, perhaps should also address the possibility of co-infections that can be occurring in moderate/severe patients, which could result in the increased CD64 expression. The authors may check the CD14 expression on neutrophils, which may partially address this.

## Reply 3

Thank you for this interesting point. As suggested, we have checked the neutrophil CD14 levels in healthy controls, moderate and severe COVID-19 (__Figure S5__). The CD14 expression was found to the highest in moderate COVID-19, yet since the expression was not controlled by isotype staining, the results may be interpreted with caution. The lowered neutrophil CD14 levels in severe COVID-19 may be a sign of emergency myelopoiesis and increased release of immature neutrophils into the circulation. In fact, low CD14 levels in pro-neutrophil-like cells were found in COVID-19 by Schulte-Schrepping and colleagues [@Schulte-Schrepping2020]. We discuss this phenomenon and mention the possibility of bacterial re-infections, especially in ventilated patients in the Discussion section of the revised manuscript.

## Issue 4

While HLA-DR, CD14, and CD16 labeling can be used to identify monocyte subsets, it is also known that HLA-DR expression can be downregulated during severe systemic inflammation, such as in COVID-19 infection. The main question is will the fact that the authors exclusively gated on HLA-DR^+^ cells for their analysis exclude some monocytes? A previous study (Silvin et al., 2020)  that is cited by the authors, specifically showed that HLA-DR^lo^ CD14^hi^ CD16^lo^ classical monocytes were detected in severe patients, suggesting a suppressive immune state. Do the authors see similar populations from their study? Additionally, why do non-classical monocytes gated (as shown in Figure 2A) appear to be CD14^+^ despite being previously described as CD14^-^?

## Reply 4

We also feel that this is an important point calling for clarification. Lowered HLA-DR levels in classical monocytes are indeed well documented in the literature in particular in severe COVID as compared with mild or moderate disease [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021]. Hence, to overcome the limitations of monocyte detection based solely on HLA-DR expression, we re-analyzed the study flow cytometry data. In more detail, neutrophils were defined within the Lineage^-^ gate by logical AND gating of the CD11b^+^, CD62L^+^, CD16^+^ and SSC^hi^ cells. To identify monocytes in the non-neutrophil cells Lineage^-^ cells (logical NOT gating), UMAP (uniform manifold approximation and projection) [@McInnes2018; @Becht2018] dimensionality reduction algorithm with CD14, CD15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR signals was applied. The monocyte cell cluster was identified by peaks of HLA-DR, CCR2, CX3CR1 and CD14 expression as compared with other non-neutrophil cell clusters (__Figure S1__). To gate the monocyte subsets, CD14, CD16, CCR2, CX3CR1, CD62L and CX3CR1 expression was analyzed (__Figure S2__). Hence, we are pretty confident that no monocytes were missed by our analysis strategy. Still, the most elegant and modern approach to the problem would be a consistent analysis of each sample with UMAP or FlowSOM, which was unfortunately computationally unfeasible (calculation time > 30 minutes per 2000000 event sample with the respective FlowJo plugins, total samples: `r nrow(mono_data$data_tbl)*10`).

Furthermore, we compared mean fluorescence intensity values for monocyte subset backbone markers HLA-DR, CCR2 and CX3CR1 in classical, intermediate and non-classical monocytes from healthy, moderate and severe COVID-19 donors (__Figure S4__). In general the pattern of the marker regulation in the monocyte subsets was similar in the study group. Interestingly, we could not detect any consistent downregulation of HLA-DR in classical monocytes in COVID-19 as compared with healthy controls, instead HLA-DR tended to be higher in classical monocytes in moderate COVID-19 as compares with healthy controls or severe disease (three group comparison: p = 0.10, Kruskal-Wallis test, healthy median: 54000 [IQR: 48000 - 65000], moderate COVID-19 median: 90000 [IQR: 45000 - 140000], severe COVID-19 median: 45000 [IQR: 31000 - 58000]). This apparent discrepancy in light of literature evidence may be a result of differences timing of the blood sampling and/or technical differences in cytometry staining as discussed in the Discussion section of the revised manuscript.

Interestingly, although no downregulation of classical HLA-DR in the investigated COVID-19 individuals could be observed, the participant cluster #2 identified by the SOM algorithm (__Figure 4__ and __Reply 2__) was characterized by the presence of CD86-, CD163-, CD279-expressing monocytes - a population which resembles the immature, immunoregulatory monocyte subset described by others [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021; @Bost2021; @Penttila2021]. We elaborate on that in the Discussion section of the revised manuscript.

Finally, the non-classical monocytes in samples measured in our laboratory are consistently measured CD14^low^ (please compare with [@Haschka2019]) and seem to be defined as such by Sylvin et al. [@Silvin2020]. Since we used other subset markers (CCR2, CX3CR1, HLA-DR, CD62L), we are confident, that the non-classical population corresponds to the non-classical subset described in the literature [@Wong2011].

## Issue 5

The authors used flow cytometry to identify neutrophils based on SSC and CD15 expression (Supplementary Figure S2). Technically, this is not entirely correct, as CD15 is a generic granulocyte marker and its expression level may vary in response to inflammation. CD15 is expressed by neutrophils, eosinophils, and even basophils, and their numbers may change during SARS-CoV2 infection. The authors should utilize CD11b, CD62L and CD16 (already in the panel) to identify neutrophils more specifically.

## Reply 5

We understand the issue. To address that, we defined neutrophils within the Lineage^-^ subset by logical AND gating of the the CD11b^+^, CD62L^+^, CD16^+^ and SSC^hi^ cells. Please refer to __Reply 4__ and __Figure S1__ for details.

## Issue 6

Typo in the subtitle "Alterations in meyloid leukocyte cellularity and phenotype may predict severe disease course", should be Myeloid.

## Reply 6

We apologize for the typo. The section was removed anyway from the revised manuscript (see: __Reply 2__, __Reviewer 1__)

\newpage

# Reviewer 2

We would like to thank the reviewer for the careful study of the manuscirpt and excellent feedback.

## Issue 1

The details of analysis inclusion scheme are given as supplementary Figure S1, however it would be better to use as a main figure, to follow the exact sample numbers involved in the study.

## Reply 1

We understand the point and have moved the study inclusion scheme to the main figures as suggested (__Figure 1__).

## Issue 2

There is a big age difference between moderate and severe COVID-19 patients. In this point what about the healthy individual characteristics included in the study? There is no information.

## Resply 2

We apologize for not providing the demographic and clinical information on healthy controls. In the revised manuscript, we provide the available data on healthy blood donors (__Table S1__). We agree with the Reviewer, that there are large and significant differences in age between the study groups and discuss it now as a study limitation. At the same time, the clustering results shown in __Figure 4__, clinical characteristic of the participant clusters (__Table S4__) and, especially no significant differences in median age between the clusters, suggest that age was not a relevant co-variate of the innate immune response patterns in the investigated time window of COVID-19. The age difference between moderate and severe COVID-19 patients likely pertains to the age as a known risk factor of severe disease and death. We elaborate on that in the Discussion section of the revised manuscript.

## Issue 3

In complete study data set Table S5, some of the parameters are showing heterogeneity. The patients having high/low WBC, IL-6 or ferritin could have something different?

## Reply 3

This is an interesting question. Multiple papers cited in the manuscript report on heterogeneous inflammatory response to COVID-19 involving both immune hyper-activation and/or immunosuppression [@Silvin2020; @Schulte-Schrepping2020; @Christensen2021; @Kvedaraite2021]. To investigate that in the study cohort, we subjected the participants to self-organizing map (SOM) clustering in respect to the myeloid leukocyte parameters obtained by out flow cytometry strategy. As described in more detail in __Reply 4__ to the __Reviewer 1__, we could identify four participant clusters differing in the quantity and surface phenotype of monocytes and neutrophils and tended to differ in the levels of inflammation and iron parameters (__Figure 4__ and __5__). The presence of such clusters suggests, that the immune response to the SARS-CoV-2 pathogen follows distinct patterns which may pertain to the hyper-activated, immunosuppressive and emergency myelopeiesis phenotypes described in the literature. We discuss that point in the Discussion section of the revised manuscript.

## Issue 4

It would be better to give some more details for the measurement of the laboratory inflammation markers in the text.

## Response 4

We thank the Reviewer for rasing this important point. We provide more details laboratory measurements in Methods of the revised manuscript (David: könntest Du sie bitte aussuchen? Danke!)

## Issue 5

Even the small numbers of participants are the limitation of the study, the results were given in details and presented with main and supplementary tables and figures, some of them can be reduced, however the study will shed light on future studies.

## Response 5

We thank the Reviewer for appreciation of our work! As suggested, we have simplified the analysis pipeline (see: __Reply 2__, __Reviewer 1__) and moved figures with particular significantly regulated cytometry features (CD64, CD86, neutrophil, Lineage^-^ and NLR) to Supplementary Material. In addition, the Result part of the text was shortened to meet the journal's word limit.

\newpage

# References

### 2022-02-23 manuscript/figure_chunks.Rmd
Initial commit
Post-revision analysis pipeline

```{r fig-consort, fig.width = 7.086614166, fig.height = 5.905511805, fig.cap = 'Analysis inclusion scheme.'}

paper_figures$consort$plot

```

__Figure \@ref(fig:fig-consort). Analysis inclusion scheme.__

```{r fig-cyto-screen, fig.width = 7.086614166, fig.height = 4.724409444, fig.cap = 'Screening of the flow cytometry features for the differences between healthy controls, moderate and severe COVID-19.'}

paper_figures$cyto_screen$plot

```

__Figure \@ref(fig:fig-cyto-screen). Screening of the flow cytometry features for the differences between healthy controls, moderate and severe COVID-19.__

```{r fig-subset-distribution, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Monocyte subset distribution in healthy controls, moderate and severe COVID-19.'}

paper_figures$subset_distribution$plot

```

__Figure \@ref(fig:fig-subset-distribution). Monocyte subset distribution in healthy controls, moderate and severe COVID-19.__

```{r fig-clustering, fig.width = 7.086614166, fig.height = 5.905511805, fig.cap = 'Unsupervised clustering of healthy controls and COVID-19 individuals by myeloid leukocyte cytometry features.'}

paper_figures$clustering$plot

```

__Figure \@ref(fig:fig-clustering). Unsupervised clustering of healthy controls and COVID-19 individuals by myeloid leukocyte cytometry features.__

```{r fig-clust-inflammation, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Differences in systemic inflammation status between the clusters of study participants.'}

paper_figures$clust_inflammation$plot

```

__Figure \@ref(fig:fig-clust-inflammation). Differences in systemic inflammation status between the clusters of study participants.__

### 2022-02-23 manuscript/main_manuscript_revision.Rmd
Initial commit
Post-revision analysis pipeline

---
title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in  hospitalized COVID-19 patients"
subtitle: "Manuscript file"
author: 
  - David Haschka^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]
  - Verena Petzer^[Department of Internal Medicine V, Medical University of Innsbruck, Innsbruck, Austria]
  - Francesco Burkert^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]
  - Gernot Fritsche^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]
  - Sophie Wildner^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]
  - Rosa Bellmann-Weiler^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]
  - Dominik Wolf^[Department of Internal Medicine V, Medical University of Innsbruck, Innsbruck, Austria] 
  - Piotr Tymoszuk^[Data Analytics As a Service Tirol, Innsbruck, Austria]
  - Guenter Weiss^[Department of Internal Medicine II, Medical University of Innsbruck, Innsbruck, Austria]

date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx

bibliography: cov_biblio.bib
csl: european-journal-of-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H}\floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = 'cairo_pdf')

```

\newpage

# Abstract

Severe 
<span custom-style = 'revision'>coronavirus disease 19 (COVID-19)</span> 
manifests with systemic immediate pro-inflammatory innate immune activation and altered iron turnover. As iron homeostasis, 
<span custom-style = 'revision'>differentiation and function of myeloid leukocytes</span> 
are interconnected we sought to characterize the cellularity, surface marker expression and iron transporter phenotype of neutrophils and monocyte subsets in COVID-19 patients within 72 hours from hospital admission, and how 
<span custom-style = 'revision'>these parameters</span> 
relate to infection severity and 
<span custom-style = 'inflammatory response.'>these parameters</span>

Between March and November 2020, blood leukocyte samples from 
<span custom-style = 'revision'>hospitalized</span> 
COVID-19 patients 
<span custom-style = 'revision'>(n = 48)</span> 
and healthy individuals 
<span custom-style = 'revision'>(n = 7)</span> 
were analyzed by flow cytometry enabling comparative analysis of 40 features. 
<span custom-style = 'revision'>Inflammation driven neutrophil expansion, depletion of CD16^+^ non-classical monocytes and changes in surface expression of neutrophil and monocyte CD64 and CD86 were associated with COVID-19 severity. By unsupervised self-organizing map clustering, four patterns of innate myeloid response were identified and linked to varying levels of systemic inflammation, iron witholding and the disease course.</span> These alterations of the myeloid leukocyte compartment 
<span custom-style = 'revision'>during acute COVID-19</span> 
may be hallmarks of inefficient viral control, 
<span custom-style = 'revision'>immune hyper-activation and immunosupression</span> and help at risk prediction and treatment 
<span custom-style = 'revision'>optimization.</span>

\newpage

# Introduction

Since late 2019, the 
<span custom-style = 'revision'>SARS-CoV-2 pandemic</span> 
has resulted in a global health crisis. 
<span custom-style = 'revision'>As of February 2022, more than 420 million infections and over 5.8 million COVID-19 deaths</span> 
were documented [@WHO2021]. 
COVID-19 is characterized by a heterogeneous clinical manifestation and divergent courses of the disease. This ranges from asymptomatic 
<span custom-style = 'revision'>infection, to ambulatory disease with wide-ranged flu-like and multi-organ complaints, up to severe respiratory or multi-organ condition </span>, 
requiring hospitalization and treatment at intensive care unit [@Sahanic2021; @Oberfeld2020]. 
Especially in the latter cases, an excessive over-stimulation of the immune system, likewise driven by an impaired immune control of viral replication leads to a ‘cytokine release syndrome’ with subsequent organ damage [@Riva2020; @Fajgenbaum2020]. Specifically, low levels of CD4^+^ and CD8^+^ cells, 
<span custom-style = 'revision'>impaired anti-viral type I interferon response</span> 
and polymorphisms in the genes involved in antigen presentation 
<span custom-style = 'revision'>were associated with the risk of severe COVID-19</span> [@Ellinghaus2020;@Zhang2020a;@Zhang2020c;@VanDerMade2020]. 
Furthermore, aberrant monocyte and neutrophil accumulation along with systemic inflammation 
<span custom-style = 'revision'>and altered systemic iron homeostasis have been linked to the disease severity and a poor prognosis</span> 
[@Schulte-Schrepping2020;@Bellmann-Weiler2021;@Merad2020;@Laing2020;@Lucas2020;@Christensen2021;@Silvin2020;@Penttila2021;@Vitte2020]. This pertains to the key role of 
<span custom-style = 'revision'>myeloid leukocytes, i. e. macrophages, monocytes and neutrophils in the SARS-CoV-2 pathogen control and orchestration of the inflammatory response</span> [@Merad2020;@Mueller2021;@Trombetta2021;@Vasse2021;@Christensen2021;@Bost2021;@Silvin2020].

Human blood monocytes consist of three populations: the major subset of 
<span custom-style = 'revision'>CD14^+/high^ CD16^-/low^ classical cells and substantially</span> 
less abundant CD14^+^ CD16^+^ intermediate and CD14^-/low^ CD16^+/high^ non-classical monocytes [@Wong2012a]. These subsets display differences in the pattern to activating stimuli and 
<span custom-style = 'revision'>immune effector functions</span> 
such as phagocytosis, antigen presentation and cytokine production [@Wong2012a]. Alterations in the monocyte subset 
<span custom-style = 'revision'>distribution and surface molecule repertoire<span> 
were observed in multiple inflammatory conditions including COVID-19 and were proposed to correlate with the disease course and prognosis [@Wong2012a;@Haschka2020;@Patel2017;@Gatti2020;@Penttila2021;@Silvin2020;@Geanon2021;@Schulte-Schrepping2020;@Trombetta2021;@Mueller2021;@Zingaropoli2021;@Christensen2021;@Vasse2021;@Lucas2020].

From the clinical and immunological point of view there is an ongoing search for markers 
<span custom-style = 'revision'>characterizing</span> the clinical course of 
<span custom-style = 'revision'>COVID-19 as well as the protective and pathological immune response to the pathogen.</span>
<span custom-style = 'revision'>Herein we investigated the surface molecule repertoire of myeloid innate immune cells from hospitalized COVID-19 patients by flow cytometry and linked the surface phenotype to the disease severity and patterns of systemic inflammation and iron turnover by self-organizing map clustering[@Kohonen1995; @Vesanto2000].</span>

\newpage

# Results

## Study cohort

The study was conducted between March and November 2020 and <span custom-style = 'revision'>encompassed healthy controls (n = 7)</span> and hospitalized COVID-19 patients (n = 53 enrolled) whose blood myeloid compartment was analyzed by flow cytometry within 72 hours from hospital admission <span custom-style = 'revision'>(__Supplementary Figure S1__ and __S2__)</span>. 
<span custom-style = 'revision'>Due to insufficient flow cytometry sample quality, 5 COVID-19 participants were excluded from the analysis. Finally, samples from `r count(suppl_tables$study_dataset, sev_paper)$n[1]` healthy controls, 
`r count(suppl_tables$study_dataset, sev_paper)$n[2]` moderate COVID-19 patients not requiring oxygen therapy during hospitalization (WHO score 3) and `r count(suppl_tables$study_dataset, sev_paper)$n[3]` severe COVID-19 individuals requiring oxygen or mechanical ventilation (WHO score 4 - 6) were available for the analysis (__Figure \@ref(fig:fig-consort)__). Among severe COVID-19 cases, `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), icu)$txt_lab[2]` were treated at an intensive case unit and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), mortality)$txt_lab[2]` deceased during hospitalization.</span> Males constituted `r my_count(filter(cohort$analysis_tbl, sev_paper == 'moderate'), sex)$txt_lab[2]` of the moderate and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), sex)$txt_lab[2]` of the severe COVID-19 group, the sex distribution difference was not significant. 
<span custom-style = 'revision'>The median age was significantly different between the healthy controls (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'healthy'), 'age')` years), moderate (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'moderate'), 'age')` years) and severe COVID-19 patients (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'severe'), 'age')`, years) and peaked in the severe COVID-19 group (__Supplementary Table S1__).</span>

## Sustained systemic and cellular inflammatory response and iron restriction in COVID-19

Systemic inflammation, as evidenced by elevated levels of interleukin-6 (IL6), C-reactive protein (CRP) and neopterin, 
<span custom-style = 'revision'>is present in hospitalized COVID-19 individuals and was</span> linked to a poor outcome [@Laing2020; @Lucas2020; @Christensen2021; @Bellmann-Weiler2021]. In our study cohort, <span custom-style = 'revision'>blood IL6, CRP and neopterin concentrations at admission were significantly higher in patients developing severe COVID-19 than in subjects with moderate disease as described previously [@Bellmann-Weiler2021] (__Supplementary Figure S3A__).</span> 
<span custom-style = 'revision'>These differences in inflammatory markers was paralleled by significantly reduced circulating iron concentrations, transferrin saturation and elevated circulating ferritin levels in severe COVID-19 as compared with the moderate group (__Supplementary Figure S3B__). Such phenomenon reflects iron restriction in response to a severe infection, reported for COVID-19 as well [@Weiss2018;@Yagc2021; @Christensen2021; @Brandtner2020; @Bellmann-Weiler2020; @Zhou2020].</span>

## Alterations of the myeloid compartment composition and surface phenotype in healthy volunteers, moderate and severe COVID-19 patients

Next, we investigated how this systemic inflammatory and <span custom-style = 'revision'>iron-restricted</span> milieu shapes the composition, <span custom-style = 'revision'>inflammatory and iron transporter protein repertoire</span> of myeloid leukocytes [@Recalcati2012; @Nairz2017]. 
<span custom-style = 'revision'>To this end, we analyzed the whole blood monocyte and neutrophil compartment of healthy controls and COVID-19 patients within 72 hours post hospital admission with multi-color flow cytometry (__Supplementary Table S2__). Neutrophils were defined within the lymphocyte lineage-negative leukocyte population (Lin^-^: CD3^-^ CD19^-^ CD56^-^) by logical AND gating of the CD16^+^, CD11b^+^, CD62L^+^ and SSC^hi^ cells. Monocytes were identified within the non-neutrophil gate by UMAP clustering (uniform manifold approximation and projection) [@McInnes2018] in respect to CD11b, CD14, CD15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression levels (__Supplementary Figure S1__). Classical, intermediate and non-classical monocyte subsets were defined by differences in CD14, CD16, HLA-DR, CCR2 and CX3CR1 surface levels [@Wong2012a] within the monocyte cluster cells (__Supplementary Figure S2__). Of note, the markers used to define monocyte populations: HLA-DR, CCR2 and CX3CR1 displayed similar expression differences between classical, intermediate and non-classical cells from healthy donors and COVID-19 patients. No downregulation of HLA-DR levels in classical monocytes described previously [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021] could be observed in COVID-19 patients as compared with healthy controls (__Supplementary Figure S4__). In turn, neutrophil CD14 expression was found significantly different between the study groups and was the highest in moderate COVID-19 individuals.</span>

<span custom-style = 'revision'>A total of `r length(cyto_clust$cyto_variables)` features could be measured with our flow cytometry pipeline including the percentages of neutrophils, classical, intermediate and non-classical monocytes as well as isotype staining-controlled surface levels of pro- and anti-inflammatory markers, iron-import protein CD71 (transferrin receptor 1) and iron exporter FPN1 (ferroportin) (__Supplementary Table S3__). Among them, `r length(cyto$kruskal_signif_features)` variables displayed significant differences between healthy controls, moderate and severe COVID-19 cell donors as determined by Kruskal-Wallis test: (1) percentage of lineage-negative cells, (2) neutrophil: lymphocyte ratio (NLR), (3) percentages of non-classical monocytes within the CD45^+^ compartment and (4) pan-monocytes, expression levels of CD64 in (5) neutrophils, (6) classical, (7) intermediate and (8) non-classical monocytes as well as surface CD86 levels in (9) classical and intermediate monocytes (10) (__Figure \@ref(fig:fig-cyto-screen)__).</span>

<span custom-style = 'revision'>More specifically, an expansion of neutrophils at the cost of Lineage-positive lymphocytes could be observed in severe COVID-19 resulting in substantially higher NLR values (median: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'nlr')`) as compared with moderate COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'nlr')`) or healthy controls (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'nlr')`) (__Supplementary Figure S6__). This phenomenon was accompanied by a depletion of the non-classical monocyte subset, which was evident in moderate COVID (median % of pan-monocytes: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) and significant in the severe disease (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) as compared with healthy cell donors (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_nonclassicalMono')`). In addition, the intermediate monocyte compartment was substantially, yet not significantly larger in COVID-19 patients (median, % of CD45^+^ cells, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_intermediateMono')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'panMono_intermediateMono')`) than in control individuals (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_intermediateMono')`) (__Figure \@ref(fig:fig-subset-distribution)__). Of note, neutrophil expansion [@Silvin2020; @Penttila2021; @Vitte2020; @Laing2020; @Christensen2021] along with a profound reduction of non-classical monocyte levels [@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020] was described as a characteristic of acute COVID-19.</span>

The neutrophil CD64 was virtually absent in healthy controls (median $\Delta$MFI: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'neutrophils_CD64')`) but highly expressed in moderate (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'neutrophils_CD64')`) and severe COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'neutrophils_CD64')`). 
A similar pattern of CD64 regulation was discerned for classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD64')`) and intermediate monocytes (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD64')`), which, however, expressed this FC$\gamma$ receptor at high levels also in non-infected controls <span custom-style = 'revision'>(_Supplementary Figure S7__)</span>.
Surface expression of CD86 in classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD86')`) and intermediate monocytes (healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD86')`) reached the maximum in moderate COVID-19 patients as compared with healthy controls and severe COVID-19 patients <span custom-style = 'revision'>(__Supplementary Figure S8__)</span>.

## Patterns of innate myloid leukocyte response in acute COVID-19

<span custom-style = 'revision'>
In order to elucidate possible patterns of co-regulation of myeloid leukocyte quantity and surface phenotype, we subjected the set of `r length(cyto_clust$cyto_variables)` flow cytometry parameters measured in our study (__Supplementary Table S3__) and the healthy and COVID-19 study participants to self-organizing map clustering [@Kohonen1995; @Vesanto2000].
</span>

<span custom-style = 'revision'>
By this means, we could identify three clusters of the cytometry features and four clusters of the study individuals (__Supplementary Figure S9__). The cytometry feature clusters included (1) a subset consisting primarily of markers of neutrophil expansion (termed 'Neutro': Lin^-^ and neutrophil levels, neutrophil CD64, NLR and monocyte: lymphocyte ratio/MLR), (2) a cluster dominated by CD86, CD163 and FPN1 expression in neutrophils and monocyte sub-populations and monocyte subset quantity variables (termed: 'CD86/CD163') and (3) a cluster encompassing mainly D40, CD71 and CD64 levels in myeloid populations (termed 'CD40/CD64/CD71') (__Figure \@ref(fig:fig-clustering)A__).
</span>

<span custom-style = 'revision'>
Healthy controls and moderate COVID-19 cases comprised 75% of the largest participant cluster #1 and was characterized by the lowest levels CD40/CD64/CD71 parameters as well as healthy-like monocyte subset distribution, neutrophil levels and neutrophil CD64 expression (__Figure \@ref(fig:fig-clustering)AB__). Interestingly, none of the severe COVID-19 cases assigned to the cluster #1 required an ICU treatment (__Supplementary Table S4__). 
The remaining clusters consisted predominantly of severe COVID-19 patients. The hallmark of the cluster #2 was an increased expression of CD86, C163, the immunosuppresive PD-L1 molecule (CD279) and the iron exporter FPN1 in classical and intermediate monocytes as well as an expansion of the intermediate subset. The participant cluster #3 in turn was primarily characterized by an elevated neutrophil and monocyte CD64 as well as a depletion of the non-classical monocytes. Finally, leukocyte samples assigned to the cluster #4 displayed a consistent expansion of the neutrophil compartment, elevated CD64 in neutrophils and a depletion non-classical monocytes (__Figure \@ref(fig:fig-clustering)A__).
</span>

<span custom-style = 'revision'>
In a comparison restricted to the COVID-19 individuals, the participant clusters demonstrated similar age and body mass index. In turn, the cluster #1 and #4 COVID-19 individuals were predominantly male (78%) as compared with the remaining subjects. Concerning the disease course, the fraction of participants requiring respiratory support was the lowest in cluster #1 and peaked in the clusters #3 and #4. However, none of these clinical features was significantly different in the all-group and the cluster #1 versus rest comparisons (__Supplementary Table S4__).
</span>

<span custom-style = 'revision'>
Finally, we investigated the inflammatory status and iron turnover parameters in COVID-19 patients assigned to the participant clusters. Although the differences between the clusters did not reached statistical significance following multiple testing correction, a clear tendency towards increased blood levels of IL6, CRP and neopterin was observed in the clusters #2, #3 and #4 as compared with the healthy-like cluster #1 (__Figure \@ref(fig:fig-clust-inflammation)__, __Supplementary Table S4__). Accordingly, the clusters #2, #3 and #4 had the lowest levels of circulating iron and transferrin saturation values suggestive of inflammatory iron restriction [@Weiss2018]. Of interest, despite the elevated inflammatory marker levels, blood ferritin concentration in the cluster #2 tended to be lower than in the similarly inflammatory clusters #3 and #4 (__Figure \@ref(fig:fig-clust-inflammation)__, __Supplementary Table S4__).
</span>

<span custom-style = 'revision'>
Collectively, the clustering analysis results suggest the presence of qualitatively and quantitatively distinct phenotypes of innate myeloid cell response to SARS-CoV-2 infection. Such phenotypic diversity is likely associated with COVID-19 severity and partially independent of the demographic patient background.
</span>

\newpage

# Discussion

Exaggerate cytokine and innate cell response is hallmark of <span custom-style = 'revision'>severe COVID-19</span> requiring hospitalization, oxygen or ICU therapy [@Schulte-Schrepping2020; @Christensen2021; @Silvin2020; @Laing2020; @Lucas2020; @Penttila2021; @Vitte2020; @Bellmann-Weiler2021].
In agreement with previous <span custom-style = 'revision'>reports</span>, we observed elevated circulating IL6, CRP and neopterin levels and an expansion of neutrophils with a concomitant reduction in lymphocytes <span custom-style = 'revision'>in COVID-19 patients requiring respiratory support</span>. Similar to other infections [@Weiss2018; @Brandtner2020], this resulted in reduced levels of circulating iron and <span custom-style = 'revision'>elevated ferritin</span>, in line with few published reports <span custom-style = 'revision'>on COVID-19</span> [@Yagc2021; @Christensen2021; @Yagc2021; @Bellmann-Weiler2020; @Zhou2020]. 
The sole cellular iron exporter FPN1 in monocyte and macrophages is down-regulated by the hormone hepcidin and diverse <span custom-style = 'revision'>inflammatory cytokines and up-regulated by cellular iron</span> [@Weiss2018; @Haschka2019; @Haschka2021; @Nemeth2004a]. Interestingly, hypoferremia in acute COVID-19 was accompanied by only mildly elevated blood hepcidin [@Yagc2021; @Zhou2020]<span custom-style = 'revision'>, which may explain the consistently elevated FPN1 levels in the subsets of inflammatory COVID-19 patients assigned to the participant cluster #2.</span>
<span custom-style = 'revision'>In general, the changes of the cellular iron trafficking and systemic turnover in COVID-19</span> may bear relevance for the effective SARS-CoV-2 control because iron levels significantly impact on differentiation of T cells, monocytes and macrophages and also affect anti-microbial immune effector mechanisms [@Weiss2018; @Haschka2021]. <span custom-style = 'revision'>As such, these processes require further mechnistic and clinical investigations.</span>

By applying flow cytometry, we were able to characterize a palette of `r length(cyto$variable)` variables associated with the quantity, pro-, anti-inflammatory <span custom-style = 'revision'>and iron transport</span> phenotype of neutrophils, classical, intermediate and non-classical monocytes.
Among them, a consistent depletion of non-classical monocytes <span custom-style = 'revision'>in part</span> paralleled by an expansion of the intermediate subset became evident in COVID-19 patients. This phenomenon was also described by others <span custom-style = 'revision'>and suggested to aggravate with infection severity</span>[@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020]. 
<span custom-style = 'revision'>It was also proposed to predict the risk of critical disease and respiratory failure [@Silvin2020; @Vasse2021; @Mueller2021].</span> 
The CD16^+^ non-classical subset contraction was suggested to be specific for COVID-19 but not for bacterial sepsis [@Christensen2021]. 
However, <span custom-style = 'revision'>results of an experimental endotoxin challenge in humans demonstrated a general depletion of blood monocytes</span> and a step-wise reconstitution involving immediately the classical population followed by the remaining subsets [@Patel2017]. Hence, the non-classical monocyte contraction in COVID-19 may reflect an early step in the non-specific inflammatory response to SARS-CoV-2. <span custom-style = 'revision'>A subsequent re-population of the non-classical compartment by classical monocyte-derived precursors during convalescence</span> can be inferred from longitudinal COVID-19 immunophenotyping studies  [@Penttila2021; @Christensen2021; @Silvin2020].

<span custom-style = 'revision'>Our immunophenotyping results</span> revealed an up-regulation of the FC$\gamma$ receptor CD64 in neutrophils, classical and intermediate monocytes <span custom-style = 'revision'>in COVID-19 patients</span> as described recently [@Geanon2021; @Vitte2020; @Silvin2020; @Schulte-Schrepping2020; @Bourgoin2021]. 
<span custom-style = 'revision'>For neutrophils, the elevated CD64 expression together with lowered neutrophil CD14 levels observed in the study group of severe COVID-19 patients may indicate the presence of less mature pro-neutrophil-like CD64^+^ CD14^low^ cells expanding in early COVID-19 [@Schulte-Schrepping2020] and sepsis [@Meghraoui-Kheddar2020; @Sack2017] and hence represent a general reaction to a strong inflammatory stimuli triggering the emergency myelopoiesis [@Schulte-Schrepping2020; @Silvin2020]. This is further supported by our clustering data, where the high neutrophil CD64 levels co-occurred with the neutrophil expansion in the highly inflammatory, severe COVID-19 dominated participant cluster #4. Since the CD64 expression in neutrophils and monocytes was the highest in the clusters #3 and #4 comprising of predominantly ventilated severe COVID-19 patients, secondary bacterial airway co-infections may also contribute to the up-regulation of this marker.</span>
The expression of CD86 in the intermediate and classical monocyte subsets followed a distinct regulatory pattern with a peak in moderate COVID-19 which is in line with other publications [@Trombetta2021; @Laing2020].
<span custom-style = 'revision'>
In the clustering analysis of our cytometry data, monocyte CD86 expression was found co-regulated with the immunosuppressive ligand CD279 (PD-L1), the alternative macrophage activation marker CD163 and the iron exporter FPN1 and was the highest in the participant cluster #2. Collectively, this suggests, that in a fraction of COVID-19 patients, subsumed under the participant cluster #2 in the study cohort, the blood monocytes may acquire an immunoregulatory, T cell-suppressive, iron export CD86^high^ CD279^high^ FPN1^high^ phenotype. Notably, expansion of a similar HLA-DR^low^, CD86^hi^, CD163^hi^ and T cell-suppressive classical monocyte populations was described for moderate and severe COVID-19 [@Silvin2020; @Penttila2021; @Christensen2021; @Schulte-Schrepping2020; @Bost2021]. In our cohort, however, no significant differences in monocyte HLA-DR expression between COVID-19 patients and healthy controls could be observed, which may be attributed to the differences in the detection methodology (multi-color flow cytometry versus CyTOF or spectral flow cytometry).
</span>

<span custom-style = 'revision'>
To get a thorough overview of myeloid leukocyte response in acute COVID-19, we performed a clustering analysis with the self-organizing map technique [@Kohonen1995; @Vesanto2000] and could identify four distinct clusters of the study participants. The largest, healthy-like cluster #1 was associated with near normal composition and surface phenotype of neutrophils and monocytes as well as the lowest levels of systemic inflammation, largely normal iron homeostasis and low risk of ICU admission. the other clusters were characterized by a clearly inflammatory and iron-restrictive background but differed in the myeloid leukocyte phenotype. In particular, the cluster #2 was hallmarked by the presence of CD86-, CD163-, CD279- and FPN1-expressing monocytes suggestive of an immunoregulatory phenotype [@Silvin2020; @Penttila2021; @Christensen2021; @Schulte-Schrepping2020; @Bost2021] as discussed above. The cluster #3 was characterized primarily by high levels of monocyte CD40 and CD64 which may reflect a pro-inflammatory priming of those cells as observed in chronic inflammatory conditions [@Luo2018; @Yang2016; @Hamers2019]. Finally, the expansion of CD64-expressing neutrophils in the participant cluster #4 may represent an inflammation-triggered emergency myelopoiesis and occurrence of immature neutrophils in the circulation of severe COVID-19 patients [@Schulte-Schrepping2020; @Silvin2020; @Meghraoui-Kheddar2020; @Sack2017]. The prognostic and therapeutic relevance of such individual heterogeneity of the innate response to the SARS-CoV-2 pathogen needs to be investigated in a larger study collective.
</span>

Our study bears limitations referring to a small number of participants which, together with the lacking verification cohort, precluded more systematic <span custom-style = 'revision'>analyses adjusted for known risk factors of severe COVID-19 such as age, sex and co-morbidity.</span>
The alterations in neutrophil and monocytes subset quantities and phenotype as well as the clinical utility of the predictive myeloid score needs to be verified in a larger follow-up study possibly including longitudinal determinations to capture dynamics of the myeloid leukocyte compartments [@Penttila2021; @Christensen2021; @Silvin2020; @Trombetta2021; @Lucas2020]. 
<span custom-style = 'revision'>Additionally, differences in age-structure and likely in frequency of co-morbidity between the healthy controls COVID-19 patients in the study cohort may have confounded the presented analysis results.</span>"

Our study corroborates the massive changes in the neutrophil and monocyte compartment in acute moderate and severe COVID-19 including expansion of neutrophils, depletion of the non-classical monocytes 
<span custom-style = 'revision'>as well as regulation of CD64 and CD86 expression in myeloid leukocytes. Finally, we demonstrate the presence of four patterns of the innate myeloid response to SARS-CoV-2 with possible implication for diagnosis, risk assessment and therapy of COVID-19.</span>.

\newpage

# Methods {#sec:methods}

## Ethics and study approval

The study performed in accordance with the Declaration of Helsinki and the European Data Policy and approved by the institutional review board of the University of Innsbruck (approval number: 1091/2020). All study participants were conscious and gave a written informed consent prior to enrollment. The participants’ data were anonymized after recording hence making identification of the particular subjects impossible.

## Study population and procedures

he study was performed between 21.03.2020 and 16.11.2020 at the hospital of the University of Innsbruck, Tyrol, Austria. The inclusion criteria for COVID-19 participants were age $\geq$ 18 years and hospital admission because of acute SARS-CoV-2 infection confirmed by nasal or oral swab PCR test. The exclusion criterion was critical physical or mental state at admission precluding giving an informed study participation consent. Healthy controls were recruited among the institution’s health care workers.

Measurement of the laboratory inflammation markers: C-reactive protein (CRP), interleukin-6 (IL6), neopterin, and iron status parameters: ferritin, iron concentration and transferrin saturation (TF-Sat), in blood at admission was done as routine determination at the hospital central laboratory. Whole venous EDTA blood (5 - 10 ml) for the flow cytometry measurement was withdrawn within 72 hours from hospital admission.

## Flow cytometry

The flow cytometry staining was done with the antibodies against the qualitative backbone markers of the lymphocyte, neutrophil and monocyte lineages and the test antibodies for qualitative markers of pro- and anti-inflammatory surface phenotype (CD40, CD64, CD80, CD86, CD163, CD274/PD-L1 and CD279/PD-1) and iron turnover proteins (CD71/transferrin receptor 1 and FPN1/ferroportin)[@Haschka2019; @Ross2017] listed in __Supplementary Table S2__. The antibodies were purchased from Beckton-Dickinson (Franklin Lakes, NJ), Thermo Fisher (Waltham. MA) or Biolegend (San Diego, CA).

Whole blood was subjected to RBC depletion with ACK (Ammonium-Chloride-Potassium) buffer, followed by extensive washing with FACS buffer (1% fetal calf serum, 0.5 mM EDTA in PBS) and staining with the backbone antibody mix (dilutions: 1:20 - 1:200 in BD Horizon™ Brilliant Stain Buffer, Beckton-Dickinson) for 10 minutes at 4 ^o^C. Next, the sample was washed, split and stained with the separate isotype or test antibody mixes (1:200 dilution in BD Horizon™ Brilliant Stain Buffer) for additional 10 minutes at 4 ^o^C. The samples were measured with a Cytoflex S device (Beckman Coulter, Brea, CA) and analyzed with FlowJo 10.7 software (Beckton-Dickinson).

The cytometry data was analyzed by an experienced operator blinded to the study group assignment with a strategy depicted in __Supplementary Figures S1__ and __S2__. 
<span custom-style = 'revision'>Neutrophils were identified within the lineage-negative leukocytes (CD3^-^ CD19^-^ CD56^-^) by logical AND gating of CD11b^+^, CD16^+^, CD62L^+^ and SSC^lo^ events. Monocytes were defined within the non-neutrophil gate (logical NOT gating) by UMAP [@McInnes2018] in respect to CD11b, CD14, C15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression. The monocyte cluster was identified by high CD14, CCR2, CX3CR1 and HLA-DR levels. Classical, intermediate and non-classical monocytes were identified within the monocyte UMAP cluster by differences in CD14, CD16, CCR2, CX3CR1, CD62L and HLA-DR expression [@Wong2012a].</span>
Surface expression was expressed as a difference in median signal intensity ($\Delta$MFI) between the specific staining and the isotype. Neutrophil CD71 and CD163 as well as CD80 staining in neutrophils and monocytes were excluded from the downstream analysis due to low and virtually invariant expression levels. Monocyte: lymphocyte ratio was calculated as a ratio of the CD45^+^ percents of HLA-DR^+^ pan-monocytes to Lineage^+^ (CD3, CD19, CD56) cells. Neutrophil: lymphocyte ratio was calculated as a ratio of the CD45^+^ percents of CD15^+^ SSC^hi^ neutrophils to Lineage^+^ cells.

## Statistical analysis

Statistical analysis was done with R version 4.0.5 and packages, _tidyverse_, <span custom-style = 'revision'>_DescTools_, _rstatix_</span> and _cowplot_ packages [@Wickham2019; @Wilke2019], <span custom-style = 'revision'>the in-house-developed package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA) and a script with clustering function wrappers (https://github.com/PiotrTymoszuk/clustering-tools-2).</span>

Since most of the analyzed variables demonstrated non-normal distribution (Shapiro-Wilk test), non-parametric hypothesis tests were utilized in the downstream analyses. To search for cytometry features (__Supplementary Table S3__) significantly differing between 
<span custom-style = 'revision'>
healthy controls. moderate and severe COVID-19 patients, Kruskal-Wallis test with 
</span>
with Benjamini-Hochberg (FDR: False Discovery Rate) correction for multiple comparisons was applied [@Benjamini1995]. 
<span custom-style = 'revision'>
The significant variables were defined by the pFDR < 0.05 cutoff. Post-hoc testing for pairwise differences between the groups was accomplished by Benjamini-Hochberg corrected Mann-Whitney U test.
</span>
<span custom-style = 'revision'>
Clustering of the flow cytometry features (__Supplementary Table S3__) and of the healthy and COVID-19 study participants in respect to the flow cytometry features was done with a combined self-organizing map (SOM) and hierarchical clustering (HCl) procedure [@Kohonen1995; @Vesanto2000]. The data were pre-processed by Z-score normalization and the participants with missing values eliminated. The cytometry features or participants were subjected to dimensionality reduction by SOM (5 $\times$ 5 hexagonal grid, cosine distance between the features/participants) and, subsequently, the SOM nodes were clustered with the HCl (cosine distance between the nodes, Ward D2 algorithm). The choice of the SOM node cluster number was motivated by the bend of the within-cluster sum-of-square curve and a visual analysis of the dendrograms (__Supplementary Figure S9__). The fraction of clustering variance (ratio of the total within-cluster to total sum of squares) was `r signif(cyto_clust$ft_diagnostic$variance$frac_var, 2)` and `r signif(cyto_clust$part_diagnostic$variance$frac_var, 2)` for the feature and participant clustering, respectively.
</span>

\newpage

# Data availability

The data set with variables analyzed in the study is available as __Supplementary Table S5__. The R analysis pipeline is available at https://github.com/PiotrTymoszuk/covFacs.

# Acknowledgments

We would like to acknowledge the suffering and commitment of the study participants and the medical staff during the COVID-19 pandemic.

# Author contribution

Conceptualization: DH, VP, DW, PT, GW; Data Curation: DH, VP; Formal Analysis: DH, PT;
Investigation: DH, VP, FB, GF, SW, RB, PT; Methodology: DH, VP, PT; Resources: DW, GW;
Visualization: DH, VP, PT; Original Draft Preparation: DH, PT; Review and Editing: DH, VP, FB, GF, SW, RB, DW, PT, GW.

# Conflict of interest

PT owns his data science enterprise, Data Analytics as a Service Tirol. PT has also received an honorarium for the study data management, curation and analysis. The author declares no other conflicts of interest. Other authors declare that non conflict of interest exists.

# Funding information

Support by the Christian Doppler Society (GW) and the Austrian Research Fund Doctoral Programm HOROS (FWF-W1253, GW, VP) and the Society for Funding Research and Education in Molecular Immunology and Infectiology ('Verein zur Förderung von Forschung und Weiterbildung in Molekularer Immunologie und Infektiologie') is gratefully acknowledged.

\newpage

# Figure Legends

__Figure \@ref(fig:fig-consort). Analysis inclusion scheme.__

__Figure \@ref(fig:fig-cyto-screen). Screening of the flow cytometry features for the differences between healthy controls, moderate and severe COVID-19.__

Levels of `r length(cyto$variable)` parameters of monocytes and neutrophils were measured by flow cytometry in healthy controls, moderate and severe COVID-19 patients within 72 hours post hospital admission (__Supplementary Table S3__). The parameter levels were compared between healthy controls, moderate and severe COVID by Kruskal-Wallis test corrected for multiple testing with Benjamini-Hochberg method (False Discovery Rate, FDR). Significance and $\eta^2$ effect size statistic is presented in the plot. Each point represents a single flow cytometry feature. Significant features are highlighted in read and labeled with their names. The horizontal dashed line represented the significance cutoff (p = 0.05). Numbers of complete observations are presented under the plot.

__Figure \@ref(fig:fig-subset-distribution). Monocyte subset distribution in healthy controls, moderate and severe COVID-19.__

Percentages of classical, intermediate and non-classical monocytes within the monocyte cluster cells (__Supplementary Figure S1__ - __S2__) were measured in healthy controls, moderate and severe COVID-19 patients. Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Representative cytometry results. Monocyte cluster cells are presented.

__(B)__ Summary plots. Kruskal-Wallis p values are indicated in the plot sub-heading, post-hoc test results are shown in the plot, numbers fo complete observations are presented next to the plot. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range.

__Figure \@ref(fig:fig-clustering). Unsupervised clustering of healthy controls and COVID-19 individuals by myeloid leukocyte cytometry features.__

Flow cytometry features (n = `r length(cyto_clust$cyto_variables)`, __Supplementary Table S3__) and study participants (n numbers presented in __B__) were subjected to clustering analysis with a combined self-organizing map (SOM) - hierarchical clustering (HCl) algorithm (__Supplementary Figure S9__). Three clusters of cytometry features termed after the most characteristic features (neutrophil, CD40, CD64, CD71, CD86 and CD163) and four participant clusters were identified.

__(A)__ Normalized levels of flow cytometry parameters (Z-score) in the participant (horizontal) and cytometry feature clusters (vertical plot facets) presented as a heat map.

__(B)__ Percentages of healthy controls, moderate and severe COVID-19 participants in the participant clusters. Statistical significance was determined by $\chi^2$ test.

__Figure \@ref(fig:fig-clust-inflammation). Differences in systemic inflammation status between the clusters of study participants.__

Blood levels of markers of inflammation (IL6: interleukin 6, C-reactive protein: CRP, neopterin) and iron turnover (ferritin, plasma iron, TF-Sat: transferrin saturation) were compared in COVID-19 patients assigned to the participant clusters (__Figure \@ref(fig:fig-clustering)A__, __Supplementary Figure S9__). Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method. Numbers of participants assigned to the clusters are indicated under the plots.

\newpage

# Figures

```{r fig-consort, fig.width = mm_inch(paper_figures$consort$w), fig.height = mm_inch(paper_figures$consort$h), fig.cap = 'Analysis inclusion scheme.'}

paper_figures$consort$plot

```

\newpage

```{r fig-cyto-screen, fig.width = mm_inch(paper_figures$cyto_screen$w), fig.height = mm_inch(paper_figures$cyto_screen$h), fig.cap = 'Screening of the flow cytometry features for the differences between healthy controls, moderate and severe COVID-19.'}

paper_figures$cyto_screen$plot

```

\newpage

```{r fig-subset-distribution, fig.width = mm_inch(paper_figures$subset_distribution$w), fig.height = mm_inch(paper_figures$subset_distribution$h), fig.cap = 'Monocyte subset distribution in healthy controls, moderate and severe COVID-19.'}

paper_figures$subset_distribution$plot

```

\newpage

```{r fig-clustering, fig.width = mm_inch(paper_figures$clustering$w), fig.height = mm_inch(paper_figures$clustering$h), fig.cap = 'Unsupervised clustering of healthy controls and COVID-19 individuals by myeloid leukocyte cytometry features.'}

paper_figures$clustering$plot

```

\newpage

```{r fig-clust-inflammation, fig.width = mm_inch(paper_figures$clust_inflammation$w), fig.height = mm_inch(paper_figures$clust_inflammation$h), fig.cap = 'Differences in systemic inflammation status between the clusters of study participants.'}

paper_figures$clust_inflammation$plot

```

\newpage

# References

### 2022-02-23 manuscript/supplement_chunks.Rmd
Initial commit
Post-revision analysis pipeline

```{r fig-gating-strategy, fig.width = 7.086614166, fig.height = 7.3228346382, fig.cap = 'Gating strategy and identification of blood neutrophils and monocytes.'}

suppl_figures$gating_strategy$plot

```

__Figure \@ref(fig:fig-gating-strategy). Gating strategy and identification of blood neutrophils and monocytes.__

```{r fig-monocyte-subsets, fig.width = 7.086614166, fig.height = 4.0551181061, fig.cap = 'Identification of blood monocyte subsets.'}

suppl_figures$monocyte_subsets$plot

```

__Figure \@ref(fig:fig-monocyte-subsets). Identification of blood monocyte subsets.__

```{r fig-systemic-inflammation, fig.width = 7.086614166, fig.height = 5.511811018, fig.cap = 'Systemic inflammation and iron turnover markers in hospitalized COVID-19 subjects.'}

suppl_figures$systemic_inflammation$plot

```

__Figure \@ref(fig:fig-systemic-inflammation). Systemic inflammation and iron turnover markers in hospitalized COVID-19 subjects.__

```{r fig-backbone, fig.width = 7.086614166, fig.height = 7.086614166, fig.cap = 'Regulation of the monocyte subset markers HLA-DR, CCR2 and CX3CR1 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$backbone$plot

```

__Figure \@ref(fig:fig-backbone). Regulation of the monocyte subset markers HLA-DR, CCR2 and CX3CR1 in healthy controls, moderate and severe COVID-19.__

```{r fig-cd14, fig.width = 7.086614166, fig.height = 3.543307083, fig.cap = 'Regulation of neutrophil CD14 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$cd14$plot

```

__Figure \@ref(fig:fig-cd14). Regulation of neutrophil CD14 in healthy controls, moderate and severe COVID-19.__

```{r fig-cell-inflammation, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Cytometry markers of myeloid leukocyte expansion in healthy controls, moderate and severe COVID-19.'}

suppl_figures$cell_inflammation$plot

```

__Figure \@ref(fig:fig-cell-inflammation). Cytometry markers of myeloid leukocyte expansion in healthy controls, moderate and severe COVID-19.__

```{r fig-myeloid-cd64, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Regulation of myeloid leukocyte CD64 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$myeloid_cd64$plot

```

__Figure \@ref(fig:fig-myeloid-cd64). Regulation of myeloid leukocyte CD64 in healthy controls, moderate and severe COVID-19.__

```{r fig-myeloid-cd86, fig.width = 7.086614166, fig.height = 5.7086614115, fig.cap = 'Regulation of monocyte CD86 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$myeloid_cd86$plot

```

__Figure \@ref(fig:fig-myeloid-cd86). Regulation of monocyte CD86 in healthy controls, moderate and severe COVID-19.__

```{r fig-cluster-qc, fig.width = 7.086614166, fig.height = 8.661417314, fig.cap = 'Training and clusterin of self-organizing maps.'}

suppl_figures$cluster_qc$plot

```

__Figure \@ref(fig:fig-cluster-qc). Training and clusterin of self-organizing maps.__

### 2022-02-23 manuscript/supplementary_material_revision.Rmd
Initial commit
Post-revision analysis pipeline

---
title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
subtitle: "Supplementary Material"
author: ""
date: "`r format(Sys.time(), '%Y-%m-%d')`"

output: 
  bookdown::word_document2:   
    reference_docx: ms_template.docx
bibliography: cov_biblio.bib
csl: european-journal-of-immunology.csl
header-includes:
  \usepackage{longtable}
  \usepackage{tabu}
  \usepackage{caption}
  \usepackage{makecell}
  \usepackage{pdflscape}
  \usepackage{array}
  \usepackage{booktabs}
  \usepackage{threeparttable}
  \usepackage{threeparttablex}
  \usepackage{wrapfig}
  \usepackage{multirow}
  \usepackage[normalem]{ulem}
  \usepackage{colortbl}
  \usepackage{xcolor}
  \usepackage{float} \floatplacement{figure}{H}\floatplacement{table}{H}
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{S\arabic{table}}       \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
---

```{r, setup, include = FALSE}

library(bookdown)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, dev = 'cairo_pdf')

```

\newpage

\beginsupplement

# Supplementary Tables {#sec:tables}

```{r tab-chohort, tab.cap = 'Characteristic of the healthy and COVID-19 study participants.'}

suppl_tables$cohort_chara %>% 
  flextable %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:4, width = 2.5, unit = 'cm') %>%
  width(5:6, width = 2.5, unit = 'cm') %>% 
  theme_vanilla() %>% 
  footnote(4, 1, value = as_paragraph('Body Mass Index'), ref_symbols = '1') %>% 
  footnote(i = c(2, 4), j = 5, value = as_paragraph('Kruskal-Wallis test'), ref_symbols = '2') %>% 
  footnote(i = c(2, 4, 5), j = 6, value = as_paragraph('Mann-Whitney test'), ref_symbols = '3') %>% 
  footnote(i = 3, j = 5:6, value = as_paragraph('\u03C7\u00B2 test'), ref_symbols = '4') %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```

\newpage

```{r tab-antibodies, tab.cap = 'Antibodies used for flow cytometry staining.'}

suppl_tables$cytom_antibodies %>% 
  flextable %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all') %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:4, width = 3, unit = 'cm') %>% 
  merge_v(1) %>% 
  theme_vanilla()

```

\newpage

```{r tab-cyto-vars, tab.cap = 'Variables obtained from cytometry staining analyzed in the study.'}

suppl_tables$cytometry_vars %>% 
  flextable %>% 
  width(1, width = 6, unit = 'cm') %>% 
  width(2, width = 3, unit = 'cm') %>% 
  theme_vanilla() %>%
  footnote(1, 2, 
           value = as_paragraph('panMono: monocyte cluster cells defined by UMAP, \u0394MFI: difference in median fluorescence intensity between the test antibody and isotype-stained sample'), 
           part = 'header') %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```

\newpage

```{r tab-cluster, tab.cap = 'Characteristic of COVID-19 patients assigned to the participant clusters defined by flow cytometry features.'}

suppl_tables$cluster_chara %>% 
  flextable %>% 
  width(1, width = 3, unit = 'cm') %>% 
  width(2:5, width = 2.3, unit = 'cm') %>%
  width(6, width = 2, unit = 'cm') %>% 
  width(7, width = 2.5, unit = 'cm') %>% 
  theme_vanilla() %>% 
  footnote(4, 1, value = as_paragraph('Body Mass Index'), ref_symbols = '1') %>% 
  footnote(i = c(2, 4, 5, 8:13), j = 6, value = as_paragraph('Kruskal-Wallis test'), ref_symbols = '2') %>% 
  footnote(i = c(2, 4, 5, 8:13), j = 7, value = as_paragraph('Mann-Whitney test'), ref_symbols = '3') %>% 
  footnote(i = c(3, 6, 7), j = 6:7, value = as_paragraph('\u03C7\u00B2 test'), ref_symbols = '4') %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```

\newpage

```{r tab-data, tab.cap = 'Complete study dataset. The table is available as a supplementary Excel file.'}

tibble(' ' = '') %>% 
  flextable()

```

\newpage

# Supplementary Figures {#sec:figures}

```{r fig-gating-strategy, fig.width = mm_inch(suppl_figures$gating_strategy$w), fig.height = mm_inch(suppl_figures$gating_strategy$h), fig.cap = 'Gating strategy and identification of blood neutrophils and monocytes.'}

suppl_figures$gating_strategy$plot

```

__Supplementary Figure S\@ref(fig:fig-gating-strategy). Gating strategy and identification of blood neutrophils and monocytes.__

Lin: lineage staining (CD3, CD19, CD56). Neutrophils were identified within the CD45^+^ Lin^-^ blood leukocyte subset by logical gating (AND) of CD16^+^, CD62L^+^, CD11b^+^ and SSC^hi^ events. To identify monocytes, the non-neutrophil cells (NOT logical gate) were subjected to UMAP (uniform manifold approximation and projection, euclidean distance, k = 9 nearest neighbors, distance cutoff = 0.5) in respect to HLA-DR, CD11b, CD14, CD16, CCR2, CX3CR1, CD62L and CD15 signals. The monocyte cluster was distinguished by high expression of HLA-DR, CD14, CCR2 and CX3CR1.

\newpage

```{r fig-monocyte-subsets, fig.width = mm_inch(suppl_figures$monocyte_subsets$w), fig.height = mm_inch(suppl_figures$monocyte_subsets$h), fig.cap = 'Identification of blood monocyte subsets.'}

suppl_figures$monocyte_subsets$plot

```

__Supplementary Figure S\@ref(fig:fig-monocyte-subsets). Identification of blood monocyte subsets.__

Monocyte cluster cells were identified as presented in __Supplementary Figure S\@ref(fig:fig-gating-strategy)__. Classical monocytes were defined as CD14^hi^ CD16^-/lo^ CCR2^hi^ HLA-DR^+^ CX3CR1^lo^ monocyte cluster cells. Intermediate monocytes were defined as CD14^int/hi^ CD16^+^ CCR2^int^ HLA-DR^bright^ CX3CR1^lo^ monocyte cluster cells. Non-classical monocytes were defined as CD14^lo^ Cd16^+^ CCR2^low^ HLA-DR^+^ CX3CR1^+^ monocyte cluster cells.

\newpage

```{r fig-systemic-inflammation, fig.width = mm_inch(suppl_figures$systemic_inflammation$w), fig.height = mm_inch(suppl_figures$systemic_inflammation$h), fig.cap = 'Systemic inflammation and iron turnover markers in hospitalized COVID-19 subjects.'}

suppl_figures$systemic_inflammation$plot

```

__Supplementary Figure S\@ref(fig:fig-systemic-inflammation). Systemic inflammation and iron turnover markers in hospitalized COVID-19 subjects.__

Markers of systemic inflammation (__A__): C-reactive protein (CRP), interleukin-6 (IL6), neopterin, and iron turnover (__B__): ferritin, iron and transferrin saturation (TF-Sat), were determined in serum of moderate and severe COVID-19 study participants at hospital admission. Statistical significance was determined by Mann-Whitney U test with Benjamini-Hochberg adjustment for multiple testing. P values are indicated in the plot sub-heading, numbers of complete observations are presented next to the plots. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range. Blue dashed lines represent the normal range of the parameter.

\newpage

```{r fig-backbone, fig.width = mm_inch(suppl_figures$backbone$w), fig.height = mm_inch(suppl_figures$backbone$h), fig.cap = 'Regulation of the monocyte subset markers HLA-DR, CCR2 and CX3CR1 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$backbone$plot

```

__Supplementary Figure S\@ref(fig:fig-backbone). Regulation of the monocyte subset markers HLA-DR, CCR2 and CX3CR1 in healthy controls, moderate and severe COVID-19.__

Surface expression of HLA-DR (__A__), CCR2 (__B__) and CX3CR1 (__C__) in classical, intermediate and non-classical monocytes (__Supplementary Figure S\@ref(fig:fig-monocyte-subsets)__) was measured as mean fluorescence intensity (MFI) in healthy controls, moderate and severe COVID-19 patients. Statistical significance of the expression differences between the monocyte subsets was determined with Friedman test (grouping factor: cell donor) with Benjamini-Hochberg adjustment for multiple testing. P values are indicated in the plot sub-heading, numbers of cell donors are presented next to the plots. Each point represents a single observation, gray lines connect values obtained from the same cell donor.

\newpage

```{r fig-cd14, fig.width = mm_inch(suppl_figures$cd14$w), fig.height = mm_inch(suppl_figures$cd14$h), fig.cap = 'Regulation of neutrophil CD14 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$cd14$plot

```

__Supplementary Figure S\@ref(fig:fig-cd14). Regulation of neutrophil CD14 in healthy controls, moderate and severe COVID-19.__

Surface expression of neutrophil CD14 (__Supplementary Figure S\@ref(fig:fig-gating-strategy)__) was measured as mean fluorescence intensity (MFI) in healthy controls, moderate and severe COVID-19 patients. Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method. Kruskal-Wallis p values are indicated in the plot sub-heading, post-hoc test results are shown in the plot, numbers of complete observations are presented next to the plot. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range.

\newpage

```{r fig-cell-inflammation, fig.width = mm_inch(suppl_figures$cell_inflammation$w), fig.height = mm_inch(suppl_figures$cell_inflammation$h), fig.cap = 'Cytometry markers of myeloid leukocyte expansion in healthy controls, moderate and severe COVID-19.'}

suppl_figures$cell_inflammation$plot

```

__Supplementary Figure S\@ref(fig:fig-cell-inflammation). Cytometry markers of myeloid leukocyte expansion in healthy controls, moderate and severe COVID-19.__

Percentages of lineage-negative cells (Lin^-^) and neutrophils within the CD45^+^ leukocyte compartment and neutrophil:leukocyte ratio (__Supplementary Figure S\@ref(fig:fig-gating-strategy)__) were measured in healthy controls, moderate and severe COVID-19 patients. Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Representative cytometry results. CD45^+^ cells are presented.

__(B)__ Summary plots. Kruskal-Wallis p values are indicated in the plot sub-heading, post-hoc test results are shown in the plot, numbers of complete observations are presented next to the plot. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range.

\newpage

```{r fig-myeloid-cd64, fig.width = mm_inch(suppl_figures$myeloid_cd64$w), fig.height = mm_inch(suppl_figures$myeloid_cd64$h), fig.cap = 'Regulation of myeloid leukocyte CD64 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$myeloid_cd64$plot

```

__Supplementary Figure S\@ref(fig:fig-myeloid-cd64). Regulation of myeloid leukocyte CD64 in healthy controls, moderate and severe COVID-19.__

Surface expression of CD64 was measured as delta median fluorescence intensity ($\Delta$MFI) versus isotype staining in neutrophils, classical and intermediate monocytes (__Supplementary Figure S\@ref(fig:fig-gating-strategy)__ - __S\@ref(fig:fig-monocyte-subsets)__) in healthy controls, moderate and severe COVID-19 patients. Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Representative cytometry results. Open histograms: isotype, tinted histograms: specific staining.

__(B)__ Summary plots. Kruskal-Wallis p values are indicated in the plot sub-heading, post-hoc test results are shown in the plot, numbers of complete observations are presented next to the plot. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range.

\newpage

```{r fig-myeloid-cd86, fig.width = mm_inch(suppl_figures$myeloid_cd86$w), fig.height = mm_inch(suppl_figures$myeloid_cd86$h), fig.cap = 'Regulation of monocyte CD86 in healthy controls, moderate and severe COVID-19.'}

suppl_figures$myeloid_cd86$plot

```

__Supplementary Figure S\@ref(fig:fig-myeloid-cd86). Regulation of monocyte CD86 in healthy controls, moderate and severe COVID-19.__

Surface expression of CD86 was measured as delta median fluorescence intensity ($\Delta$MFI) versus isotype staining in classical and intermediate monocytes (__Supplementary Figure S\@ref(fig:fig-monocyte-subsets)__) in healthy controls, moderate and severe COVID-19 patients. Statistical significance was determined by Kruskal-Wallis test with Mann-Whitney post-hoc test. Testing results were adjusted for multiple comparisons with Benjamini-Hochberg method.

__(A)__ Representative cytometry results. Open histograms: isotype, tinted histograms: specific staining.

__(B)__ Summary plots. Kruskal-Wallis p values are indicated in the plot sub-heading, post-hoc test results are shown in the plot, numbers of complete observations are presented next to the plot. Each point represents a single observation, boxes represent medians with interquartile range (IQR), whiskers span over the 150% IQR range.

\newpage

```{r fig-cluster-qc, fig.width = mm_inch(suppl_figures$cluster_qc$w), fig.height = mm_inch(suppl_figures$cluster_qc$h), fig.cap = 'Training and clustering of self-organizing maps.'}

suppl_figures$cluster_qc$plot

```

__Supplementary Figure S\@ref(fig:fig-cluster-qc). Training and clustering of self-organizing maps.__

Flow cytometry parameters (__Supplementary Table S\@ref(tab:tab-cyto-vars)__) and study participants were subjected to self-organizing map (SOM) dimensionality reduction (5 $\times$ 5 hexagonal grid, cosine distance between the observations) followed by SOM node clustering with Ward D2 algorithm (cosine distance between the nodes).

__(A)__ SOM training process for the cytometry parameters and study participants. Mean distance to the SOM winning unit as a function of algorithm iteration is presented. Each point represents a single iteration blue lines depict LOESS (locally weighted scatterplot smoothing) trends.

__(B)__ Clustering of the SOM nodes for the cytometry parameters and study participants. Total within cluster sum-of-squares are shown as a function of cluster numbers. Red lines indicate the selected optimal number of clusters.

__(C)__ Clustering of the SOM nodes for the cytometry parameters and study participants. Node clustering dendrograms are presented.

### 2022-02-23 analysis scripts/cluster_characteristic.R

@@ -1,0 +1,217 @@
+# Compares demographic, inflammatory and iron paramaters in the clusters ----
+
+  insert_head()
+
+# container list -----
+
+  clust_chara <- list()
+  
+# globals -----
+  
+  clust_chara$infl_variable <- c('il6', 'crp', 'neopterin', 
+                                 'ferritin', 'iron', 'tf_sat')
+  
+  clust_chara$variable <- c('age', 'sex', 'bmi', 'hospital_stay', 'oxygen', 'icu',
+                            clust_chara$infl_variable)
+  
+  clust_chara$types <- c('kruskal_test', 'chisq_test', 'kruskal_test', 'kruskal_test', 'chisq_test', 'chisq_test', 
+                         rep('kruskal_test', 6))
+  
+  clust_chara$post_hoc_types <- c('wilcoxon_r', 'cramer_v', 'wilcoxon_r', 'wilcoxon_r', 'cramer_v', 'cramer_v', 
+                                  rep('wilcoxon_r', 6))
+  
+  ## analysis table
+  
+  clust_chara$analysis_tbl <- extract(cyto_clust$part_object, 'assignment') %>% 
+    mutate(id = observation) %>% 
+    left_join(mono_data$data_tbl[c('id', clust_chara$variable, 'sev_paper')], by = 'id')
+  
+  ## n numbers
+  
+  clust_chara$n_numbers <- clust_chara$analysis_tbl %>% 
+    count(clust_id)
+  
+  
+  ## participant cluster colors
+  
+  clust_chara$clust_colors <- c('#1' = 'darkolivegreen3', 
+                                '#2' = 'cornflowerblue', 
+                                '#3' = 'coral3', 
+                                '#4' = 'gray60')
+  
+# Normality and EOV of the numeric features ----
+  
+  insert_msg('Normality and EOV of the numeric variables')
+  
+  clust_chara$normality <- explore(filter(clust_chara$analysis_tbl, 
+                                          sev_paper %in% c('moderate', 'severe')), 
+                                   variables = c('age', 'bmi', 'hospital_stay', 
+                                                 'il6', 'crp', 'neopterin', 
+                                                 'ferritin', 'iron', 'tf_sat'), 
+                                   what = 'normality', 
+                                   pub_styled = TRUE)
+  
+  clust_chara$eov <- compare_variables(filter(clust_chara$analysis_tbl, 
+                                              sev_paper %in% c('moderate', 'severe')), 
[…]

### 2022-02-23 analysis scripts/clustering.R

@@ -1,0 +1,137 @@
+# Clusters participant in respect to the cytometry features
+
+  insert_head()
+  
+# container list -----
+  
+  cyto_clust <- list()
+  
+# globals: variables, normalized analysis table -----
+  
+  insert_msg('Globals setup')
+
+  ## cytometry features
+  
+  cyto_clust$cyto_variables <- proj_globals$var_tbl %>% 
+    filter(type == 'response', 
+           description == 'cytomtery') %>% 
+    .$variable
+  
+  cyto_clust$var_labs <- translate_var(cyto_clust$cyto_variables, type = 'plot_lab') %>% 
+    stri_replace(fixed = ' percent', replacement = 's') %>% 
+    stri_replace(fixed = 'monocyte', replacement = 'mono') %>% 
+    stri_replace(fixed = 'Monocyte', replacement = 'Mono') %>% 
+    stri_replace(fixed = 'Monos', replacement = 'Mono') %>% 
+    stri_replace(fixed = 'monos', replacement = 'mono') %>% 
+    stri_replace(fixed = 'Classical', replacement = 'Class.') %>% 
+    stri_replace(fixed = 'classical', replacement = 'Class.') %>% 
+    stri_replace(fixed = 'Intermediate', replacement = 'Int.') %>% 
+    stri_replace(fixed = 'Mono: Lymphocyte Ratio', replacement = 'MLR') %>% 
+    stri_replace(fixed = 'Neutrophil: Lymphocyte Ratio', replacement = 'NLR') %>% 
+    stri_replace(fixed = 'of ', replacement = '') %>% 
+    stri_replace(regex = '^\\s{1}', replacement = '') %>% 
+    stri_replace(fixed = ', \u0394MFI', replacement = '') %>% 
+    set_names(cyto_clust$cyto_variables)
+  
+  ## clustering table
+  
+  cyto_clust$analysis_tbl <- mono_data$data_tbl %>% 
+    select(id, all_of(cyto_clust$cyto_variables)) %>% 
+    filter(complete.cases(.)) %>% 
+    map_dfc(function(x) if(is.numeric(x)) scale(x)[, 1] else x) %>% 
+    column_to_rownames('id')
+
+# clustering of the cytometry features ------
+  
+  insert_msg('Clustering of the cytometry features')
+  
+  cyto_clust$ft_object <- combi_cluster(data = t(cyto_clust$analysis_tbl), 
+                                        distance_som = 'cosine', 
+                                        xdim = 4, 
+                                        ydim = 5 , 
[…]

### 2022-02-23 analysis scripts/correlations.R

@@ -1,0 +1,86 @@
+# This script investigates correlations between the CD64, Cd40, Cd86 and monocyte subset levels
+# and CRP and IL-6 in the CoV strata
+
+  insert_head()
+  
+# tools -----
+  
+  library(cowplot)
+  
+# data container -----
+  
+  cyto_corr <- list()
+  
+# globals -----
+  
+  insert_msg('Globals setup')
+  
+  ## variables of interest, significant features identified by serial
+  ## testing of the cytometry features with Kruskal test
+
+  cyto_corr$cyto_var <- cyto$kruskal_signif_features
+  
+  cyto_corr$labor_vars <- c('crp', 'il6', 'neopterin', 'iron', 'ferritin')
+  
+  cyto_corr$pairs <- cyto_corr$cyto_var %>% 
+    map(function(cyto_var) cyto_corr$labor_vars %>% 
+          map(~c(cyto_var, .x))) %>% 
+    unlist(recursive = FALSE)
+  
+  cyto_corr$pairs <- cyto_corr$pairs %>% 
+    map(paste, collapse = '_') %>% 
+    set_names(cyto_corr$pairs, .)
+
+# Serial correlation with Spearman method (non-normal distribution of the features) -----
+  
+  insert_msg('Serial correlation analysis')
+  
+  cyto_corr$test_results <- cyto_corr$pairs %>% 
+    map_dfr(~correlate_variables(filter(mono_data$data_tbl[.x], 
+                                        complete.cases(mono_data$data_tbl[.x])), 
+                                 variables = .x, 
+                                 what = 'correlation', 
+                                 type = 'spearman', 
+                                 pub_styled = TRUE)) %>% 
+    mutate(corr_id = names(cyto_corr$pairs), 
+           p_adjusted = p.adjust(p_value, 'BH'), 
+           significance = ifelse(p_adjusted < 0.001, 
+                                 'p < 0.001', 
+                                 ifelse(p_adjusted >= 0.05, 
+                                        paste0('ns (p = ', signif(p_adjusted, 2), ')'), 
+                                        paste('p =', signif(p_adjusted, 2)))), 
+           plot_cap = paste(eff_size, significance, sep = ', '), 
+           plot_cap = stri_replace(plot_cap, fixed = 'rho', replacement = '\u03C1'))
+  
+# Plotting the correlations ----
+  
+  insert_msg('Plotting the correlations')
+  
[…]

### 2022-02-23 analysis scripts/cytometry.R

@@ -1,0 +1,208 @@
+# Compares the cytometry paramater values between healthy, moderate and severe CoV.
+# Kruskal-Wallis test with eta-squared as an effect size statistic (eta < 0 are considered 0, i.e. no effect)
+# Post-hoc test: Wilcoxon with the r effect size statistic
+
+  insert_head()
+  
+# container list ------
+  
+  cyto <- list()
+  
+# globals -----
+  
+  insert_msg('Globals setup')
+  
+  ## variables
+  
+  cyto$variable <- proj_globals$var_tbl %>% 
+    filter(type == 'response', 
+           description == 'cytomtery') %>% 
+    .$variable
+  
+  ## variable labels
+  
+  cyto$var_labs <- translate_var(cyto$variable, type = 'plot_lab') %>% 
+    stri_replace(fixed = ' percent', replacement = 's') %>% 
+    stri_replace(fixed = 'monocyte', replacement = 'mono') %>% 
+    stri_replace(fixed = 'Monocyte', replacement = 'Mono') %>% 
+    stri_replace(fixed = 'Monos', replacement = 'Mono') %>% 
+    stri_replace(fixed = 'monos', replacement = 'mono') %>% 
+    stri_replace(fixed = 'Classical', replacement = 'Class.') %>% 
+    stri_replace(fixed = 'classical', replacement = 'Class.') %>% 
+    stri_replace(fixed = 'Intermediate', replacement = 'Int.') %>% 
+    stri_replace(fixed = 'Mono: Lymphocyte Ratio', replacement = 'MLR') %>% 
+    stri_replace(fixed = 'Neutrophil: Lymphocyte Ratio', replacement = 'NLR') %>% 
+    stri_replace(fixed = 'of ', replacement = '') %>% 
+    stri_replace(regex = '^\\s{1}', replacement = '') %>% 
+    stri_replace(fixed = ', \u0394MFI', replacement = '') %>% 
+    set_names(cyto$variable)
+  
+  ## pairs for comparisons
+  
+  cyto$pairs <- list(healthy_moderate = c('healthy', 'moderate'), 
+                     healthy_severe = c('healthy', 'severe'), 
+                     moderate_severe = c('moderate', 'severe'))
+  
+# Normality and variance homogeneity check -------
+  
+  insert_msg('Normality and eov')
+  
+  cyto$normality <- explore(data = mono_data$data_tbl, 
+                            variables = cyto$variable, 
[…]

### 2022-02-23 analysis scripts/cytometry_backbone.R

@@ -1,0 +1,169 @@
+# Compares expression of the selected/reviewer-requested backbone cytometry features
+# between the monocyte populations and disease severity strata
+
+  insert_head()
+  
+# container list -----
+  
+  back <- list()
+  
+# globals setup ------
+  
+  ## backbone markars
+  
+  back$mono_markers <- c('ccr2', 
+                         'cd62l', 
+                         'hladr', 
+                         'cx3cr1')
+  
+  back$marker_labs <- c(ccr2 = 'CCR2', 
+                        cd62l = 'CD62L', 
+                        hladr = 'HLA-DR', 
+                        cx3cr1 = 'CX3CR1')
+  
+  ## monocvyte population colors and labels
+  
+  back$mono_colors <- c(classical = 'firebrick4', 
+                        intermediate = 'gray60', 
+                        nonclassical = 'steelblue3')
+  
+  back$mono_labs <- c(classical = 'Classical', 
+                      intermediate = 'Intermediate', 
+                      nonclassical = 'Non-classical')
+  
+  ## analysis table
+  
+  back$analysis_tbl <- back$mono_markers %>% 
+    map(~select(mono_data$data_tbl, id, sev_paper, ends_with(.x))) %>% 
+    map(~filter(.x, complete.cases(.x)))
+  
+  back$analysis_tbl <- map2(back$analysis_tbl, 
+                            back$mono_markers, 
+                            ~gather(.x, 
+                                    key = 'population', 
+                                    value = !!.y, 
+                                    ends_with(.y))) %>% 
+    map(mutate, 
+        population = stri_extract(population, 
+                                  regex = 'classical|intermediate|nonclassical'), 
+        population = factor(population, c('classical', 'intermediate', 'nonclassical'))) %>% 
+    reduce(left_join, by = c('id', 'sev_paper', 'population'))
+  
+# Normality and EOV check -----
+  
+  insert_msg('Normality and EOV check')
+  
+  back$normality <- back$analysis_tbl %>% 
+    ddply(.(sev_paper, population), 
+          explore, 
+          variables = back$mono_markers, 
+          what = 'normality', 
+          pub_styled = TRUE) %>% 
+    as_tibble
+  
+
+  back$eov_population <- back$analysis_tbl %>% 
+    ddply(.(sev_paper), 
+          compare_variables, 
+          variables = back$mono_markers, 
+          split_factor = 'population', 
+          what = 'variance', 
+          pub_styled = TRUE) %>% 
+    as_tibble
+  
[…]

### 2022-02-23 analysis scripts/cytometry_cd14.R

@@ -1,0 +1,93 @@
+# Checks differences in CD14 expression between the severity groups
+
+  insert_head()
+  
+# container list -----
+  
+  cd14 <- list()
+  
+# globals -----
+  
+  insert_msg('Globals setup')
+  
+  cd14$pairs <- list(healthy_moderate = c('healthy', 'moderate'), 
+                     healthy_severe = c('healthy', 'severe'), 
+                     moderate_severe = c('moderate', 'severe'))
+  
+# normality and EOV check -----
+  
+  insert_msg('Normality and EOV check')
+  
+  cd14$normality <- explore(mono_data$data_tbl, 
+                            variables = 'neutrophils_cd14', 
+                            what = 'normality', 
+                            pub_styled = TRUE)
+  
+  cd14$eov <- compare_variables(mono_data$data_tbl, 
+                                variables = 'neutrophils_cd14', 
+                                split_factor = 'sev_paper', 
+                                what = 'variance', 
+                                pub_styled = TRUE)
+  
+# descriptive statistic -----
+  
+  insert_msg('Descriptive statistic')
+  
+  cd14$desc_stats <- explore(mono_data$data_tbl, 
+                             variables = 'neutrophils_cd14', 
+                             split_factor = 'sev_paper', 
+                             what = 'table', 
+                             pub_styled = TRUE)
+  
+# Kruskal-Wallis test and post-hoc testing with Wilcoxon -----
+  
+  insert_msg('Kruskal-Wallis with Wilcoxon post-hoc test')
+  
+  cd14$kruskal_results <- compare_variables(mono_data$data_tbl, 
+                                            variables = 'neutrophils_cd14', 
+                                            split_factor = 'sev_paper', 
+                                            what = 'test', 
+                                            types = 'kruskal_test', 
+                                            ci = FALSE, 
+                                            pub_styled = FALSE)
+  
+  cd14$post_hoc_results <- cd14$pairs %>% 
+    map(~filter(mono_data$data_tbl, sev_paper %in% .x)) %>% 
+    map(~compare_variables(.x, 
+                           variables = 'neutrophils_cd14', 
+                           split_factor = 'sev_paper', 
+                           what = 'eff_size', 
+                           types = 'wilcoxon_r', 
+                           ci = FALSE, 
+                           pub_styled = FALSE))
+  
+# Plotting: box plot ------
+  
+  insert_msg('Box plot with the results')
+  
[…]

### 2022-02-23 analysis scripts/inflammatory_params.R

@@ -1,0 +1,117 @@
+# This script analyzes a set of inflammatory parameters in the COVID-19 severity groups: moderate vs severe/critical
+# Healthy and non-COVID-19 cases are ignored for now
+
+  insert_head()
+  
+# data container -----
+  
+  inflammation <- list()
+  
+# script globals -----
+  
+  insert_msg('Globals setup')
+  
+  ## inflammatory variables
+  
+  inflammation$variable <- c('wbc', 
+                             'crp',
+                             'pct', 
+                             'il6', 
+                             'neopterin', 
+                             'ferritin',
+                             'iron', 
+                             'tf_sat')
+  
+  ## normal ranges of the variables
+  
+  inflammation$lower_limit <- list('wbc' = 4.5, 
+                                   'crp' = 0,
+                                   'pct' = 0, 
+                                   'il6' = 0, 
+                                   'neopterin' = 0, 
+                                   'ferritin' = 12,
+                                   'iron' = 10.74, 
+                                   'tf_sat' = 15)
+  
+  inflammation$upper_limit <- list('wbc' = 11, 
+                                   'crp' = 3,
+                                   'pct' = 0.05, 
+                                   'il6' = 7, 
+                                   'neopterin' = 10, 
+                                   'ferritin' = 300,
+                                   'iron' = 30, 
+                                   'tf_sat' = 50)
+
+# analysis table: admission, CoV  and healthy cases only ------
+  
+  insert_msg('Analysis table')
+  
+  inflammation$analysis_tbl <- mono_data$data_tbl %>% 
+    filter(covid_status  == 'COVID-19') %>% 
+    select(id, 
+           sev_paper, 
+           all_of(inflammation$variable))
+  
+# variable distribution, normality test (Shapiro) -----
+  
+  insert_msg('Normality check')
+  
+  inflammation$normality <- explore(data = inflammation$analysis_tbl, 
+                                    variables = inflammation$variable, 
+                                    what = 'normality')
+  
+# Descriptive statistic ------
+  
+  insert_msg('Descriptive statistics in the severity groups')
+  
+  inflammation$desc_stats <- explore(data = inflammation$analysis_tbl, 
+                                     split_factor = 'sev_paper', 
+                                     variables = inflammation$variable, 
+                                     what = 'table')
+  
+# Comparing the values between the groups -----
+  # the variables are in substantial part non-normally distributed, hence Wilcoxon test
+  
+  insert_msg('Comparing the variables between the severity strata')
+  
[…]

### 2022-02-23 data_import.R

@@ -1,0 +1,148 @@
+# This script reads and clears the cytometry, sample assignment and clinics data
+
+
+# toolbox -----
+
+  library(soucer)
+  library(plyr)
+  library(tidyverse)
+  library(readxl)
+  library(stringi)
+
+  c('globals_setup.R', 
+    './tools/project_tools.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+  insert_head()
+
+# data container ----
+
+  mono_data <- list()
+  
+# reading the file manifest, filtering out the samples with bad quality and follow-ups ------
+  
+  insert_msg('Reading the file manifest/analysis log')
+  
+  mono_data$analysis_log <- read_csv('./cytometry/analysis_log.csv') %>% 
+    mutate(data_folder = paste('./cytometry/', sample, sep = '')) %>% 
+    filter(quality != 'bad', 
+           !stri_detect(sample, fixed = 'fup'))
+
+# reading the cytometry data: frequency, CD274/CD279 expression and the test antibody panel signals -----
+
+  insert_msg('reading the cytometry data')
+  
+  if(!'base_cyto.csv' %in% list.files('./cytometry')) {
+    
+    mono_data$base_cyto <- list(paste0(mono_data$analysis_log$data_folder, '/frequency_revision.txt'), 
+                                paste0(mono_data$analysis_log$data_folder, '/pd_revision.txt'), 
+                                paste0(mono_data$analysis_log$data_folder, '/testAb_revision.txt'), 
+                                mono_data$analysis_log$data_folder) %>% 
+      map2(., list(read_freq_data, 
+                   read_pd_data, 
+                   read_testab_data, 
+                   read_revision), 
+           ~.y(.x))
+    
+    mono_data$base_cyto <- mono_data$base_cyto %>% 
+      reduce(full_join, by = 'id')
+    
+    write_tsv(mono_data$base_cyto, './cytometry/base_cyto.csv')
+    
+  } else {
+    
+    mono_data$base_cyto <- read_tsv('./cytometry/base_cyto.csv')
+    
+  }
+
+# reading and clearing the sample assignment and clinical data ------
+  
+  insert_msg('reading the sample assigment table')
+  
+  mono_data$clinics <- read_excel('./clinics/clinical_data.xlsx')
+  
+  ## re-coding the COVID-19 status and disease severity
+  
+  mono_data$clinics <- mono_data$clinics %>% 
+    mutate(covid_status = ifelse(infection %in% c('other', 'pulmonary'), 
[…]

### 2022-02-23 exec_all.R

@@ -1,0 +1,17 @@
+# This program exectutes the KIAA project scripts
+
+# libraries -----
+
+  library(soucer)
+
+# executing the scripts ----
+
+  exec_log <- source_all(c('data_import.R', 
+                           'paper_analyses.R', 
+                           'render_paper.R'), 
+                         message = TRUE, 
+                         crash = FALSE)
+  
+  print(exec_log)
+  
+# END -----

### 2022-02-23 figure and table scripts/cohort_characteristic.R

@@ -1,0 +1,129 @@
+# This script analyzes the features in the moderate and severe CoV and makes a raw cohort characteristic table
+
+  insert_head()
+  
+# data container ----
+  
+  cohort <- list()
+
+# globals and analysis table -----
+  
+  insert_msg('Globals and analysis table')
+  
+  ## variables of interest
+  
+  cohort$variable <- c('age', 
+                       'sex', 
+                       'bmi',  
+                       'hospital_stay',
+                       'oxygen', 
+                       'icu', 
+                       'mortality')
+
+  ## healthy control data
+  
+  cohort$healthy_data <- read_excel('./clinics/healthy_controls.xlsx') %>% 
+    mutate(oxygen = 'no', 
+           icu = 'no', 
+           hospital_stay = 0, 
+           mortality = 'no', 
+           sev_paper = 'healthy')
+  
+  ## analysis table
+  
+  cohort$analysis_tbl <- mono_data$data_tbl %>% 
+    filter(covid_status == 'COVID-19') %>% 
+    select(id, 
+           sev_paper, 
+           all_of(cohort$variable)) %>% 
+    rbind(cohort$healthy_data, .) %>% 
+    mutate(icu = factor(icu, c('no', 'yes')), 
+           oxygen = factor(oxygen, c('no', 'yes')), 
+           mortality = factor(mortality, c('no', 'yes')), 
+           sex = factor(sex, c('female', 'male')))
+  
+  ## n numbers
+  
+  cohort$n_numbers <- cohort$analysis_tbl %>% 
+    count(sev_paper)
+  
+# Distribution and EOV of the numeric variables -----
+  
+  insert_msg('Normality and EOV of the numeric variables')
+  
+  cohort$normality <- explore(cohort$analysis_tbl, 
+                              variables = c('age', 'bmi', 'hospital_stay'), 
+                              what = 'normality', 
+                              pub_styled = TRUE)
+  
+  cohort$eov <- compare_variables(cohort$analysis_tbl, 
+                                  variables = c('age', 'bmi', 'hospital_stay'), 
+                                  split_factor = 'sev_paper', 
+                                  what = 'variance', pub_styled = TRUE)
+  
+# Descriptive statistic -----
+  
+  insert_msg('Descriptive statistics')
+  
+  cohort$desc_stats <- explore(cohort$analysis_tbl, 
+                               variables = cohort$variable, 
+                               split_factor = 'sev_paper', 
+                               what = 'table', 
+                               pub_styled = TRUE) %>% 
[…]

### 2022-02-23 figure and table scripts/deploy_paper.R

@@ -1,0 +1,70 @@
+# this script renders word files with the main manuscript and supplements
+
+  insert_head()
+  
+# Exporting the figure and supplementary figure code chunks ----
+  
+  insert_msg('Inserting the figures and supplementary figures')
+  
+  insert_figure(paper_figures$consort, 
+                paper_figures$cyto_screen, 
+                paper_figures$subset_distribution, 
+                paper_figures$clustering, 
+                paper_figures$clust_inflammation, 
+                file = './manuscript/figure_chunks.Rmd', 
+                ref_names = stri_replace_all(names(paper_figures), fixed = '_', replacement = '-'), 
+                captions = c('Analysis inclusion scheme.', 
+                             'Screening of the flow cytometry features for the differences between healthy controls, moderate and severe COVID-19.', 
+                             'Monocyte subset distribution in healthy controls, moderate and severe COVID-19.', 
+                             'Unsupervised clustering of healthy controls and COVID-19 individuals by myeloid leukocyte cytometry features.', 
+                             'Differences in systemic inflammation status between the clusters of study participants.'))
+  
+  insert_figure(suppl_figures$gating_strategy, 
+                suppl_figures$monocyte_subsets, 
+                suppl_figures$systemic_inflammation, 
+                suppl_figures$backbone, 
+                suppl_figures$cd14, 
+                suppl_figures$cell_inflammation, 
+                suppl_figures$myeloid_cd64, 
+                suppl_figures$myeloid_cd86, 
+                suppl_figures$cluster_qc, 
+                file = './manuscript/supplement_chunks.Rmd', 
+                ref_names = stri_replace_all(names(suppl_figures), fixed = '_', replacement = '-'), 
+                captions = c('Gating strategy and identification of blood neutrophils and monocytes.', 
+                             'Identification of blood monocyte subsets.', 
+                             'Systemic inflammation and iron turnover markers in hospitalized COVID-19 subjects.', 
+                             'Regulation of the monocyte subset markers HLA-DR, CCR2 and CX3CR1 in healthy controls, moderate and severe COVID-19.', 
+                             'Regulation of neutrophil CD14 in healthy controls, moderate and severe COVID-19.', 
+                             'Cytometry markers of myeloid leukocyte expansion in healthy controls, moderate and severe COVID-19.', 
+                             'Regulation of myeloid leukocyte CD64 in healthy controls, moderate and severe COVID-19.', 
[…]

### 2022-02-23 figure and table scripts/paper_figures.R

@@ -1,0 +1,321 @@
+# This script generates figures and supplementary figures for the paper
+
+  insert_head()
+
+# data containers -----
+  
+  paper_figures <- list()
+  suppl_figures <- list()
+  
+# Figure 1: CONSORT ------
+  
+  insert_msg('Figure 1: CONSORT')
+  
+  paper_figures$consort <- ggdraw() + 
+    draw_image('./aux figure files/consort.png')
+  
+  paper_figures$consort <- paper_figures$consort %>% 
+    as_figure(label = 'figure_1_consort', 
+              w = 180, 
+              h = 150)
+  
+# Figure 2: cytometry features differentiating between healthy, moderate and severe COVID-19 -----
+  
+  insert_msg('Figure 2: cytometry features differentiating between healthy, moderate and severe COVID-19')
+  
+  paper_figures$cyto_screen <- cyto$kruskal_plot + 
+    labs(y = expression('Kruskal test: -log'[10]*' pFDR'), 
+         x = expression('Effect size: '*eta^2))
+  
+  paper_figures$cyto_screen <- paper_figures$cyto_screen %>% 
+    as_figure(label = 'figure_2_cyto_screen', 
+              w = 180, 
+              h = 120)
+
+# Figure 3: monocyte subset distribution in healthy and CoV strata ------
+  
+  insert_msg('Figure 3: Monocyte subset distribution')
+  
+  paper_figures$subset_distribution <- make_cyto_panel(plot_list = cyto$box_plots[c('panMono_classicalMono', 
+                                                                                    'panMono_intermediateMono', 
+                                                                                    'panMono_nonclassicalMono')], 
+                                                       represent_path = './representative data/monocyte subsets/subsets_representative.png') %>% 
+    as_figure(label = 'figure_3_monocyte_subsets', 
+              w = 180, 
+              h = 220)
+
+# Figure 4: clustering of the cytometry features and study participants -----
+  
+  insert_msg('Figure 4: clustering of the participants')
+  
+  paper_figures$clustering <- plot_grid(cyto_clust$heat_map + 
+                                          theme(plot.tag = element_blank(), 
+                                                legend.position = 'bottom'), 
+                                        plot_grid(clust_chara$severity_plot + 
+                                                    theme(legend.position = 'bottom'), 
+                                                  ggdraw(), 
+                                                  nrow = 2, 
+                                                  rel_heights = c(0.7, 0.3)), 
[…]

### 2022-02-23 figure and table scripts/paper_tables.R

@@ -1,0 +1,96 @@
+# This script prepares three supplementary tables for the manuscript:
+# cohort characteristic, cytometry staining antibodies, cytometry variables analyzed in the study
+# and with all data records of the study for data distribution
+
+  insert_head()
+  
+# data container -----
+  
+  suppl_tables <- list()
+  
+# cohort characteristic ------
+  
+  insert_msg('Table S1: Cohort characteristic')
+  
+  suppl_tables$cohort_chara <- cohort$result_tbl %>% 
+    set_names(c('Variable', 
+                'Healthy', 
+                'Moderate COVID-19', 
+                'Severe COVID-19', 
+                'Comparison: all groups', 
+                'Comparison: COVID-19'))
+  
+# cytometry antibodies -----
+  
+  insert_msg('Table S2: Cytometry antibodies')
+  
+  suppl_tables$cytom_antibodies <- read_excel('./clinics/antibodies.xlsx') %>% 
+    set_names(c('Staining type', 
+                'Antigen', 
+                'Clone', 
+                'Fluorophore'))
+  
+# cytometry variables analyzed in the study -----
+  
+  insert_msg('Table S3: cytometry variables')
+  
+  suppl_tables$cytometry_vars <- proj_globals$var_tbl %>% 
+    filter(type != 'ignore', 
+           description == 'cytomtery') %>% 
+    .$variable
+  
+  suppl_tables$cytometry_vars <- tibble(Variable = translate_var(suppl_tables$cytometry_vars, type = 'label'), 
+                                        Unit = translate_var(suppl_tables$cytometry_vars, type = 'unit'))
+
+# characteristic of the participant clusters ------
+  
+  insert_msg('Clinical characteristic of the participant clusters')
+  
+  suppl_tables$cluster_chara <- clust_chara$result_tbl %>% 
+    set_names(c('Variable', 
+                'Cluster #1', 
+                'Cluster #2', 
+                'Cluster #3', 
+                'Cluster #4', 
+                'Comparison: all groups', 
+                'Comaprison: Cluster #1'))
+  
+# study data -----
+  
+  insert_msg('Table S4: study data set')
+  
+  suppl_tables$study_variables <- c(cohort$variable, 
+                                    inflammation$variable, 
+                                    cyto$variable) %>% 
+    unique
+  
+  suppl_tables$study_dataset <- mono_data$data_tbl %>% 
[…]

### 2022-02-23 globals_setup.R

@@ -1,0 +1,99 @@
+# This script provides globals for plotting and analysis
+
+  insert_head()
+  
+# toolbox -----
+  
+  library(tidyverse)
+  library(readxl)
+
+# data container -----
+  
+  proj_globals <- list()
+  
+# general graphics theme -----
+  
+  insert_msg('Theme setup')
+  
+  proj_globals$common_text <- element_text(size = 8, 
+                                           face = 'plain', 
+                                           color = 'black')
+  
+  proj_globals$common_margin <- ggplot2::margin(t = 5, 
+                                                l = 5, 
+                                                r = 5, 
+                                                b = 5, 
+                                                unit = 'mm')
+  
+  proj_globals$common_theme <- theme_classic() + theme(axis.text = proj_globals$common_text, 
+                                                       axis.title = proj_globals$common_text, 
+                                                       plot.title = element_text(size = 9, 
+                                                                                 face = 'bold'), 
+                                                       plot.subtitle = proj_globals$common_text, 
+                                                       plot.tag = element_text(size = 8, 
+                                                                               face = 'plain', 
+                                                                               color = 'black', 
+                                                                               hjust = 0, 
+                                                                               vjust = 1), 
+                                                       plot.tag.position = 'bottom', 
+                                                       legend.text = proj_globals$common_text, 
+                                                       legend.title = proj_globals$common_text, 
+                                                       strip.text = proj_globals$common_text,
+                                                       plot.margin = proj_globals$common_margin, 
+                                                       panel.grid.major = element_line(color = 'gray90'))
+  
+# severity and COVID-19 colors ------
+  
+  insert_msg('Color setup')
+  
+  proj_globals$severity_colors <- c(healthy = 'cornflowerblue', 
+                                    moderate = 'cornsilk', 
+                                    severe = 'coral2', 
+                                    critical = 'firebrick4')
+  
+  proj_globals$severity_labels <- c(healthy = 'Healthy', 
+                                    moderate = 'Moderate', 
+                                    severe = 'Severe', 
+                                    critical = 'Critical')
+  
+  proj_globals$covid_labels <- c('healthy' = 'Healthy', 
[…]

### 2022-02-23 paper_analyses.R

@@ -1,0 +1,29 @@
+# This is a mother script executing the paper analyses ------
+
+# tools -----
+
+  insert_head()
+
+  library(exda)
+
+  c('globals_setup.R', 
+    './tools/project_tools.R', 
+    './tools/clust_tools2.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+# executable scripts ------
+  
+  insert_msg('Executing paper analysis scripts')
+  
+  c('./analysis scripts/inflammatory_params.R', ## comparison of the inflammatory and iron parameters
+    './analysis scripts/cytometry.R', ## test cytometry features
+    './analysis scripts/cytometry_cd14.R', ## CD14 in neutrophils, as requested by the reviewer
+    './analysis scripts/cytometry_backbone.R', ## monocyte backbone markers as requested by the reviewer
+    './analysis scripts/correlations.R', ## correlations between the cytometry features and inflammatory/iron parameter
+    './analysis scripts/clustering.R', ## participant and cytometry feature clustering
+    './analysis scripts/cluster_characteristic.R') %>% ## characteristic of the clusters
+    source_all(message = TRUE, crash = TRUE)
+
+# END ----
+  
+  insert_tail()

### 2022-02-23 render_paper.R

@@ -1,0 +1,32 @@
+# This is a mother script for rendering the paper figures, supplementary figures and tables ------
+
+# tools -----
+
+  insert_head()
+
+  library(figur)
+  library(cowplot)
+  library(knitr)
+  library(rmarkdown)
+  library(bookdown)
+  library(flextable)
+  library(soucer)
+  library(writexl)
+
+  c('./tools/project_tools.R', 
+    'globals_setup.R') %>% 
+    source_all(message = TRUE, crash = TRUE)
+
+# executable scripts ------
+
+  insert_msg('Executing paper table, figure and publishing scripts')
+  
+  c('./figure and table scripts/paper_figures.R', ## figures and supplementary figures
+    './figure and table scripts/cohort_characteristic.R', ## clinical and demographic characteristic of the cohort
+    './figure and table scripts/paper_tables.R', ## supplement tables
+    './figure and table scripts/deploy_paper.R') %>% ## renders the manuscript and the supplementary material
+    source_all(message = TRUE, crash = TRUE)
+
+# END ----
+
+insert_tail()

### 2022-02-23 tools/clust_tools2.R

@@ -1,0 +1,2434 @@
+## a script providing tools for hierarchical, kmeans, PAM, DB scan and SOM clustering
+
+# tools -----
+
+  require(tidyverse)
+  require(philentropy)
+  require(factoextra)
+  require(dendextend)
+  require(nomclust)
+  require(cluster)
+  require(pcaPP)
+  require(dbscan)
+  require(rlang)
+  require(kohonen)
+  require(Rcpp)
+  require(grDevices)
+  require(cowplot)
+  require(ggrepel)
+  require(coxed)
+  require(caret)
+  require(furrr)
+
+  ## additional kernels for kohonen::som()
+
+  tryCatch(library(somKernels), 
+           error = function(e) install.packages('./tools/somKernels.tar', 
+                                                repos = NULL, 
+                                                type = 'source', 
+                                                INSTALL_opts = '--no-help'))
+
+  map <- purrr::map
+
+  map_som <- kohonen::map
+  
+# helper functions -----
+
+  calculate_dist <- function(data, method) {
+    
+    ## calculates a dissimilarity matrix
+    
+    if(!is_tibble(data) & !is.data.frame(data) & !is.matrix(data)) {
+      
+      stop('Provide a valid data.frame, tibble or matrix as input data', call. = FALSE)
+      
+    }
+    
+    av_distances <- c(getDistMethods(), 
+                      'smc', 
+                      'sumofsquares')
+    
+    if(method == 'smc') {
+      
+      dist_mtx <- data %>% 
+        as.data.frame %>% 
+        sm %>% 
+        as.matrix
+      
+    } else {
+      
+      if(method == 'sumofsquares') method <- 'squared_euclidean'
+      
+      dist_mtx <- data %>% 
+        as.matrix %>% 
+        distance(method = method, 
+                 use.row.names = TRUE, 
+                 mute.message = TRUE)
+      
+      if(method %in% c('cosine', 'ruzicka')) {
+        
+        ### handling the 0 - 1 similarity coefficients
+        
+        dist_mtx <- 1 - dist_mtx
+        
+      }
+      
+    }
+    
+    return(dist_mtx)
+    
+  }
+  
+  plot_knn_distance <- function(diss_obj, 
+                                k, 
+                                eps = NULL, 
+                                plot_title = NULL, 
+                                plot_subtitle = NULL, 
+                                plot_tag = NULL, 
+                                cust_theme = theme_classic()) {
+    
+    ## makes k-NN distance plot
+    
+    sort_distances <- kNNdist(x = diss_obj, 
[…]

### 2022-02-23 tools/project_tools.R

@@ -1,0 +1,503 @@
+# this script provides general functions for the project analyses
+
+# other tools -----
+
+  require(tidyverse)
+  require(ggpubr)
+  require(ggsignif)
+  require(soucer)
+  require(stringi)
+
+# Data import and clearing -----
+
+  read_freq_data <- function(path) {
+    
+    ## reads and clears a file with the population frequency data
+    
+    if(length(path) > 1) {
+      
+      data_lst <- path %>% 
+        map(read_freq_data)
+      
+      status <- data_lst %>% 
+        map_chr(function(x) if(is.null(x)) 'failure' else 'OK') %>% 
+        tibble(path = path, 
+               status = .) %>% 
+        filter(status == 'failure')
+      
+      if(nrow(status) == 0) {
+        
+        message('No reading failures!')
+        
+      } else {
+        
+        print(status)
+        
+      }
+      
+      return(data_lst %>% 
+               do.call('rbind', .))
+      
+      
+    }
+    
+    freq_data <- try(read_tsv(path) %>% 
+                       filter(!.data[['Sample:']] %in% c('Mean', 'SD')), 
+                     silent = T)
+    
+    if(any(class(freq_data) == 'try-error')) {
+      
+      message(paste('Reading failure: ', path))
+      
+      return(NULL)
+      
+    }
+    
+    return(freq_data %>% 
+             select(-`Sample:`))
+    
+  }
+  
+  read_pd_data <- function(path, exprs_only = TRUE) {
+    
+    ## reads and clears a file with the PD staining values
+    ## returns delta MFI (median)
+    
+    if(length(path) > 1) {
+      
+      data_lst <- path %>% 
+        map(read_pd_data)
+      
+      status <- data_lst %>% 
+        map_chr(function(x) if(is.null(x)) 'failure' else 'OK') %>% 
+        tibble(path = path, 
+               status = .) %>% 
+        filter(status == 'failure')
+      
+      if(nrow(status) == 0) {
+        
+        message('No reading failures!')
+        
+      } else {
+        
+        print(status)
+        
+      }
+      
+      return(data_lst %>% 
+               do.call('rbind', .))
+      
+      
+    }
+    
+    pd_tbl <- try(read_tsv(path)  %>% 
+                    filter(!.data[['Sample:']] %in% c('Mean', 'SD')), 
+                  silent = TRUE)
+    
+    if(any(class(pd_tbl) == 'try-error') | 
[…]

### 2022-02-28 figure and table scripts/paper_figures.R
Figure and table script patches

@@ -275,7 +275,7 @@
                                         plot(cyto_clust$part_object, 
                                              type = 'training', 
                                              cust_theme = proj_globals$common_theme)$observation + 
-                                          labs(title = 'Cytometry features: SOM training') + 
+                                          labs(title = 'Participants: SOM training') + 
                                           theme(plot.subtitle = element_blank(), 
                                                 plot.tag = element_blank()), 
                                         cyto_clust$ft_diagnostic$diagn_plots$node$wss + 

### 2022-02-28 manuscript/rebuttal_letter.Rmd
Supplementary Material and Figure patches

@@
 ---
-title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
+title: "Alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
 subtitle: "Point-to-Point Reply"

### 2022-02-28 manuscript/main_manuscript_revision.Rmd
Supplementary Material and Figure patches

@@
 ---
-title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in  hospitalized COVID-19 patients"
+title: "Alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in  hospitalized COVID-19 patients"
 subtitle: "Manuscript file"
@@
 <span custom-style = 'revision'>Our immunophenotyping results</span> revealed an up-regulation of the FC$\gamma$ receptor CD64 in neutrophils, classical and intermediate monocytes <span custom-style = 'revision'>in COVID-19 patients</span> as described recently [@Geanon2021; @Vitte2020; @Silvin2020; @Schulte-Schrepping2020; @Bourgoin2021].
-<span custom-style = 'revision'>For neutrophils, the elevated CD64 expression together with lowered neutrophil CD14 levels observed in the study group of severe COVID-19 patients may indicate the presence of less mature pro-neutrophil-like CD64^+^ CD14^low^ cells expanding in early COVID-19 [@Schulte-Schrepping2020] and sepsis [@Meghraoui-Kheddar2020; @Sack2017] and hence represent a general reaction to a strong inflammatory stimuli triggering the emergency myelopoiesis [@Schulte-Schrepping2020; @Silvin2020].
+<span custom-style = 'revision'>For neutrophils, the elevated CD64 expression together with lowered neutrophil CD14 levels observed in the study group of severe COVID-19 patients may indicate the presence of less mature pro-neutrophil-like CD64^+^ CD14^low^ cells expanding in acute COVID-19 [@Schulte-Schrepping2020] and sepsis [@Meghraoui-Kheddar2020; @Sack2017] and hence represent a general reaction to a strong inflammatory stimuli triggering the emergency myelopoiesis [@Schulte-Schrepping2020; @Silvin2020].
 This is further supported by our clustering data, where the high neutrophil CD64 levels co-occurred with the neutrophil expansion in the highly inflammatory, severe COVID-19 dominated participant cluster #4.

### 2022-02-28 manuscript/supplementary_material_revision.Rmd
Supplementary Material and Figure patches

@@
 ---
-title: "Early alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
+title: "Alterations of blood monocyte subset distribution and surface phenotype are linked to infection severity in COVID-19 inpatients"
 subtitle: "Supplementary Material"
@@
 
-Markers of systemic inflammation (__A__): C-reactive protein (CRP), interleukin-6 (IL6), neopterin, and iron turnover (__B__): ferritin, iron and transferrin saturation (TF-Sat), were determined in serum of moderate and severe COVID-19 study participants at hospital admission.
+Markers of systemic inflammation (__A__): C-reactive protein (CRP), interleukin-6 (IL6), neopterin, and iron turnover (__B__): ferritin, iron and transferrin saturation (TF-Sat), were determined in plasma of moderate and severe COVID-19 study participants at hospital admission.
 Statistical significance was determined by Mann-Whitney U test with Benjamini-Hochberg adjustment for multiple testing.

### 2022-03-25 tools/clust_tools2.R
Not needed any more

@@ -1,2434 +1,0 @@
-## a script providing tools for hierarchical, kmeans, PAM, DB scan and SOM clustering
-
-# tools -----
-
-  require(tidyverse)
-  require(philentropy)
-  require(factoextra)
-  require(dendextend)
-  require(nomclust)
-  require(cluster)
-  require(pcaPP)
-  require(dbscan)
-  require(rlang)
-  require(kohonen)
-  require(Rcpp)
-  require(grDevices)
-  require(cowplot)
-  require(ggrepel)
-  require(coxed)
-  require(caret)
-  require(furrr)
-
-  ## additional kernels for kohonen::som()
-
-  tryCatch(library(somKernels), 
-           error = function(e) install.packages('./tools/somKernels.tar', 
-                                                repos = NULL, 
-                                                type = 'source', 
-                                                INSTALL_opts = '--no-help'))
-
-  map <- purrr::map
-
-  map_som <- kohonen::map
-  
-# helper functions -----
-
-  calculate_dist <- function(data, method) {
-    
-    ## calculates a dissimilarity matrix
-    
-    if(!is_tibble(data) & !is.data.frame(data) & !is.matrix(data)) {
-      
-      stop('Provide a valid data.frame, tibble or matrix as input data', call. = FALSE)
-      
-    }
-    
-    av_distances <- c(getDistMethods(), 
-                      'smc', 
-                      'sumofsquares')
-    
-    if(method == 'smc') {
-      
-      dist_mtx <- data %>% 
-        as.data.frame %>% 
-        sm %>% 
-        as.matrix
-      
-    } else {
-      
-      if(method == 'sumofsquares') method <- 'squared_euclidean'
-      
-      dist_mtx <- data %>% 
-        as.matrix %>% 
-        distance(method = method, 
-                 use.row.names = TRUE, 
-                 mute.message = TRUE)
-      
-      if(method %in% c('cosine', 'ruzicka')) {
-        
-        ### handling the 0 - 1 similarity coefficients
-        
-        dist_mtx <- 1 - dist_mtx
-        
-      }
-      
-    }
-    
-    return(dist_mtx)
-    
-  }
-  
-  plot_knn_distance <- function(diss_obj, 
-                                k, 
-                                eps = NULL, 
-                                plot_title = NULL, 
-                                plot_subtitle = NULL, 
-                                plot_tag = NULL, 
-                                cust_theme = theme_classic()) {
-    
-    ## makes k-NN distance plot
-    
-    sort_distances <- kNNdist(x = diss_obj, 
[…]

### 2022-03-25 manuscript/main_manuscript_revision.Rmd
Pipeline update
Clustering tools provided now by a development package

@@
 <span custom-style = 'revision'>these parameters</span> 
-relate to infection severity and 
-<span custom-style = 'inflammatory response.'>these parameters</span>
+relate to infection severity.
 
@@
 <span custom-style = 'revision'>Inflammation driven neutrophil expansion, depletion of CD16^+^ non-classical monocytes and changes in surface expression of neutrophil and monocyte CD64 and CD86 were associated with COVID-19 severity.
-By unsupervised self-organizing map clustering, four patterns of innate myeloid response were identified and linked to varying levels of systemic inflammation, iron witholding and the disease course.</span> These alterations of the myeloid leukocyte compartment
+By unsupervised self-organizing map clustering, four patterns of innate myeloid response were identified and linked to varying levels of systemic inflammation, altered cellular iron trafficking and the severity of disease.</span> These alterations of the myeloid leukocyte compartment
 <span custom-style = 'revision'>during acute COVID-19</span> 
-may be hallmarks of inefficient viral control, 
-<span custom-style = 'revision'>immune hyper-activation and immunosupression</span> and help at risk prediction and treatment 
+may be hallmarks of inefficient viral control and  
+<span custom-style = 'revision'>immune hyper-activation and may</span> and help at risk prediction and treatment 
 <span custom-style = 'revision'>optimization.</span>
@@
 has resulted in a global health crisis. 
-<span custom-style = 'revision'>As of February 2022, more than 420 million infections and over 5.8 million COVID-19 deaths</span> 
+<span custom-style = 'revision'>As of March 2022, more than 472 million infections and over 6.1 million COVID-19 deaths</span> 
 were documented [@WHO2021]. 
-COVID-19 is characterized by a heterogeneous clinical manifestation and divergent courses of the disease. This ranges from asymptomatic 
-<span custom-style = 'revision'>infection, to ambulatory disease with wide-ranged flu-like and multi-organ complaints, up to severe respiratory or multi-organ condition </span>, 
-requiring hospitalization and treatment at intensive care unit [@Sahanic2021; @Oberfeld2020]. 
+COVID-19 is characterized by a heterogeneous clinical manifestation and divergent courses of the disease [@Burkert2021]. This ranges from asymptomatic 
+<span custom-style = 'revision'>infection, to ambulatory disease with wide-ranged flu-like complaints, up to severe respiratory or multi-organ failure</span>, 
+requiring hospitalization and treatment at intensive care unit [@Sahanic2021; @Oberfeld2020; @Burkert2021]. 
 Especially in the latter cases, an excessive over-stimulation of the immune system, likewise driven by an impaired immune control of viral replication leads to a ‘cytokine release syndrome’ with subsequent organ damage [@Riva2020; @Fajgenbaum2020].
 Specifically, low levels of CD4^+^ and CD8^+^ cells,
-<span custom-style = 'revision'>impaired anti-viral type I interferon response</span> 
-and polymorphisms in the genes involved in antigen presentation 
-<span custom-style = 'revision'>were associated with the risk of severe COVID-19</span> [@Ellinghaus2020;@Zhang2020a;@Zhang2020c;@VanDerMade2020]. 
+<span custom-style = 'revision'>impaired anti-viral type I interferon response  
+and polymorphisms in the genes involved in antigen presentation and immune cell communication were associated with the risk of severe COVID-19</span> [@Ellinghaus2020; @Zhang2020a; @Zhang2020c; @VanDerMade2020].
 Furthermore, aberrant monocyte and neutrophil accumulation along with systemic inflammation 
 <span custom-style = 'revision'>and altered systemic iron homeostasis have been linked to the disease severity and a poor prognosis</span> 
-[@Schulte-Schrepping2020;@Bellmann-Weiler2021;@Merad2020;@Laing2020;@Lucas2020;@Christensen2021;@Silvin2020;@Penttila2021;@Vitte2020]. This pertains to the key role of 
+[@Schulte-Schrepping2020; @Bellmann-Weiler2021; @Merad2020; @Laing2020; @Lucas2020; @Christensen2021; @Silvin2020; @Penttila2021; @Vitte2020; @Bellmann-Weiler2020]. This pertains to the key role of 
 <span custom-style = 'revision'>myeloid leukocytes, i.
 e.
-macrophages, monocytes and neutrophils in the SARS-CoV-2 pathogen control and orchestration of the inflammatory response</span> [@Merad2020;@Mueller2021;@Trombetta2021;@Vasse2021;@Christensen2021;@Bost2021;@Silvin2020].
+macrophages, monocytes and neutrophils in the SARS-CoV-2 pathogen control and orchestration of the inflammatory response</span> [@Merad2020; @Mueller2021; @Trombetta2021; @Vasse2021; @Christensen2021; @Bost2021; @Silvin2020].
+<span custom-style = 'revision'>In addition, monocytes/macrophages play an important role in the orchestration of iron homeostasis, and inflammatory processes result in macrophage/monocyte iron retention whereas iron exerts distinct effect on macrophage polarization and immune effector functions [@Haschka2021; @Recalcati2012; @Lanser2021].</span>
 
@@
 <span custom-style = 'revision'>CD14^+/high^ CD16^-/low^ classical cells and substantially</span> 
-less abundant CD14^+^ CD16^+^ intermediate and CD14^-/low^ CD16^+/high^ non-classical monocytes [@Wong2012a]. These subsets display differences in the pattern to activating stimuli and 
+less abundant CD14^+/high^ CD16^+/high^ intermediate and CD14^-/low^ CD16^+/high^ non-classical monocytes [@Wong2012a]. These subsets display differences in the pattern to activating stimuli and 
 <span custom-style = 'revision'>immune effector functions</span> 
@@
 <span custom-style = 'revision'>distribution and surface molecule repertoire<span> 
-were observed in multiple inflammatory conditions including COVID-19 and were proposed to correlate with the disease course and prognosis [@Wong2012a;@Haschka2020;@Patel2017;@Gatti2020;@Penttila2021;@Silvin2020;@Geanon2021;@Schulte-Schrepping2020;@Trombetta2021;@Mueller2021;@Zingaropoli2021;@Christensen2021;@Vasse2021;@Lucas2020].
+were observed in multiple inflammatory conditions including COVID-19 and were proposed to correlate with the disease course and prognosis [@Wong2012a; @Haschka2020; @Patel2017; @Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020].
 
@@
 <span custom-style = 'revision'>COVID-19 as well as the protective and pathological immune response to the pathogen.</span>
-<span custom-style = 'revision'>Herein we investigated the surface molecule repertoire of myeloid innate immune cells from hospitalized COVID-19 patients by flow cytometry and linked the surface phenotype to the disease severity and patterns of systemic inflammation and iron turnover by self-organizing map clustering[@Kohonen1995; @Vesanto2000].</span>
+<span custom-style = 'revision'>Herein we investigated the surface molecule repertoire of myeloid innate immune cells from hospitalized COVID-19 patients by flow cytometry and linked the surface phenotype to the disease severity and patterns of systemic inflammation and iron turnover by self-organizing map clustering [@Kohonen1995; @Vesanto2000].</span>
 
@@
 The study was conducted between March and November 2020 and <span custom-style = 'revision'>encompassed healthy controls (n = 7)</span> and hospitalized COVID-19 patients (n = 53 enrolled) whose blood myeloid compartment was analyzed by flow cytometry within 72 hours from hospital admission <span custom-style = 'revision'>(__Supplementary Figure S1__ and __S2__)</span>.
-<span custom-style = 'revision'>Due to insufficient flow cytometry sample quality, 5 COVID-19 participants were excluded from the analysis.
+<span custom-style = 'revision'>Due to insufficient flow cytometry sample quality, 5 COVID-19 patients were excluded from further analysis.
 Finally, samples from `r count(suppl_tables$study_dataset, sev_paper)$n[1]` healthy controls,
-`r count(suppl_tables$study_dataset, sev_paper)$n[2]` moderate COVID-19 patients not requiring oxygen therapy during hospitalization (WHO score 3) and `r count(suppl_tables$study_dataset, sev_paper)$n[3]` severe COVID-19 individuals requiring oxygen or mechanical ventilation (WHO score 4 - 6) were available for the analysis (__Figure \@ref(fig:fig-consort)__).
-Among severe COVID-19 cases, `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), icu)$txt_lab[2]` were treated at an intensive case unit and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), mortality)$txt_lab[2]` deceased during hospitalization.</span> Males constituted `r my_count(filter(cohort$analysis_tbl, sev_paper == 'moderate'), sex)$txt_lab[2]` of the moderate and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), sex)$txt_lab[2]` of the severe COVID-19 group, the sex distribution difference was not significant.
+`r count(suppl_tables$study_dataset, sev_paper)$n[2]` moderate COVID-19 patients not requiring supplemental oxygen therapy during hospitalization (WHO score 3) and `r count(suppl_tables$study_dataset, sev_paper)$n[3]` severe COVID-19 individuals requiring low/high flow oxygen treatment or mechanical ventilation (WHO score 4 - 6) were available for the analysis (__Figure \@ref(fig:fig-consort)__).
+Among severe COVID-19 cases, `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), icu)$txt_lab[2]` were treated at an intensive care unit and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), mortality)$txt_lab[2]` deceased during hospitalization.</span> Males constituted `r my_count(filter(cohort$analysis_tbl, sev_paper == 'moderate'), sex)$txt_lab[2]` of the moderate and `r my_count(filter(cohort$analysis_tbl, sev_paper == 'severe'), sex)$txt_lab[2]` of the severe COVID-19 group, the sex distribution difference was not significant.
 <span custom-style = 'revision'>The median age was significantly different between the healthy controls (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'healthy'), 'age')` years), moderate (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'moderate'), 'age')` years) and severe COVID-19 patients (`r my_median(filter(cohort$analysis_tbl, sev_paper == 'severe'), 'age')`, years) and peaked in the severe COVID-19 group (__Supplementary Table S1__).</span>
@@
 Systemic inflammation, as evidenced by elevated levels of interleukin-6 (IL6), C-reactive protein (CRP) and neopterin, 
-<span custom-style = 'revision'>is present in hospitalized COVID-19 individuals and was</span> linked to a poor outcome [@Laing2020; @Lucas2020; @Christensen2021; @Bellmann-Weiler2021].
-In our study cohort, <span custom-style = 'revision'>blood IL6, CRP and neopterin concentrations at admission were significantly higher in patients developing severe COVID-19 than in subjects with moderate disease as described previously [@Bellmann-Weiler2021] (__Supplementary Figure S3A__).</span>
+<span custom-style = 'revision'>was present in hospitalized COVID-19 individuals and was</span> linked to a poor outcome [@Laing2020; @Lucas2020; @Christensen2021; @Bellmann-Weiler2021].
+In our study cohort, <span custom-style = 'revision'>plasma IL6, CRP and neopterin concentrations at admission were significantly higher in patients developing severe COVID-19 than in subjects with moderate disease as described previously [@Bellmann-Weiler2021] (__Supplementary Figure S3A__).</span>
 <span custom-style = 'revision'>These differences in inflammatory markers was paralleled by significantly reduced circulating iron concentrations, transferrin saturation and elevated circulating ferritin levels in severe COVID-19 as compared with the moderate group (__Supplementary Figure S3B__).
-Such phenomenon reflects iron restriction in response to a severe infection, reported for COVID-19 as well [@Weiss2018;@Yagc2021; @Christensen2021; @Brandtner2020; @Bellmann-Weiler2020; @Zhou2020].</span>
+Such phenomenon reflects iron restriction in response to a severe infection, reported previously for COVID-19 [@Weiss2018; @Yagc2021; @Christensen2021; @Brandtner2020; @Bellmann-Weiler2020; @Zhou2020].</span>
 
@@
 Next, we investigated how this systemic inflammatory and <span custom-style = 'revision'>iron-restricted</span> milieu shapes the composition, <span custom-style = 'revision'>inflammatory and iron transporter protein repertoire</span> of myeloid leukocytes [@Recalcati2012; @Nairz2017].
-<span custom-style = 'revision'>To this end, we analyzed the whole blood monocyte and neutrophil compartment of healthy controls and COVID-19 patients within 72 hours post hospital admission with multi-color flow cytometry (__Supplementary Table S2__).
-Neutrophils were defined within the lymphocyte lineage-negative leukocyte population (Lin^-^: CD3^-^ CD19^-^ CD56^-^) by logical AND gating of the CD16^+^, CD11b^+^, CD62L^+^ and SSC^hi^ cells.
+<span custom-style = 'revision'>To this end, we analyzed the whole blood monocyte and neutrophil compartment of healthy controls and COVID-19 patients within 72 hours of hospital admission with multi-color flow cytometry (__Supplementary Table S2__).
+Neutrophils were defined within the lymphocyte lineage-negative leukocyte population (Lin^-^: CD3^-^ CD19^-^ CD56^-^) by logical gating (AND) of the CD16^+^, CD11b^+^, CD62L^+^ and SSC^hi^ cells.
 Monocytes were identified within the non-neutrophil gate by UMAP clustering (uniform manifold approximation and projection) [@McInnes2018] in respect to CD11b, CD14, CD15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression levels (__Supplementary Figure S1__).
 Classical, intermediate and non-classical monocyte subsets were defined by differences in CD14, CD16, HLA-DR, CCR2 and CX3CR1 surface levels [@Wong2012a] within the monocyte cluster cells (__Supplementary Figure S2__).
-Of note, the markers used to define monocyte populations: HLA-DR, CCR2 and CX3CR1 displayed similar expression differences between classical, intermediate and non-classical cells from healthy donors and COVID-19 patients.
-No downregulation of HLA-DR levels in classical monocytes described previously [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021] could be observed in COVID-19 patients as compared with healthy controls (__Supplementary Figure S4__).
-In turn, neutrophil CD14 expression was found significantly different between the study groups and was the highest in moderate COVID-19 individuals.</span>
+Of note, the additional markers used to define monocyte populations (HLA-DR, CCR2 and CX3CR1) displayed similar expression differences between classical, intermediate and non-classical cells from healthy donors and COVID-19 patients.
+No downregulation of HLA-DR levels in classical monocytes, described previously [@Schulte-Schrepping2020; @Silvin2020; @Christensen2021], could be observed in COVID-19 patients as compared with healthy controls (__Supplementary Figure S4__).
+In turn, neutrophil CD14 expression was found significantly different between the study groups and was the highest in moderate COVID-19 individuals (__Supplementary Figure S5__).</span>
 
 <span custom-style = 'revision'>A total of `r length(cyto_clust$cyto_variables)` features could be measured with our flow cytometry pipeline including the percentages of neutrophils, classical, intermediate and non-classical monocytes as well as isotype staining-controlled surface levels of pro- and anti-inflammatory markers, iron-import protein CD71 (transferrin receptor 1) and iron exporter FPN1 (ferroportin) (__Supplementary Table S3__).
-Among them, `r length(cyto$kruskal_signif_features)` variables displayed significant differences between healthy controls, moderate and severe COVID-19 cell donors as determined by Kruskal-Wallis test: (1) percentage of lineage-negative cells, (2) neutrophil: lymphocyte ratio (NLR), (3) percentages of non-classical monocytes within the CD45^+^ compartment and (4) pan-monocytes, expression levels of CD64 in (5) neutrophils, (6) classical, (7) intermediate and (8) non-classical monocytes as well as surface CD86 levels in (9) classical and intermediate monocytes (10) (__Figure \@ref(fig:fig-cyto-screen)__).</span>
+Among them, `r length(cyto$kruskal_signif_features)` variables displayed significant differences between healthy controls, moderate and severe COVID-19 cell donors as determined by Kruskal-Wallis test: (1) percentage of lineage-negative cells, (2) neutrophil: lymphocyte ratio (NLR), (3) percentages of non-classical monocytes within the CD45^+^ compartment and (4) pan-monocytes, expression levels of CD64 in (5) neutrophils, (6) classical, (7) intermediate and (8) non-classical monocytes as well as surface CD86 levels in (9) classical and (10) intermediate monocytes (__Figure \@ref(fig:fig-cyto-screen)__).</span>
 
@@
 This phenomenon was accompanied by a depletion of the non-classical monocyte subset, which was evident in moderate COVID (median % of pan-monocytes: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) and significant in the severe disease (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_nonclassicalMono')`) as compared with healthy cell donors (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_nonclassicalMono')`).
-In addition, the intermediate monocyte compartment was substantially, yet not significantly larger in COVID-19 patients (median, % of CD45^+^ cells, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_intermediateMono')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'panMono_intermediateMono')`) than in control individuals (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_intermediateMono')`) (__Figure \@ref(fig:fig-subset-distribution)__).
-Of note, neutrophil expansion [@Silvin2020; @Penttila2021; @Vitte2020; @Laing2020; @Christensen2021] along with a profound reduction of non-classical monocyte levels [@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020] was described as a characteristic of acute COVID-19.</span>
+In addition, the intermediate monocyte compartment was substantially, yet not significantly, larger in COVID-19 patients (median, % of CD45^+^ cells, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'panMono_intermediateMono')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'panMono_intermediateMono')`) than in control individuals (`r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'panMono_intermediateMono')`) (__Figure \@ref(fig:fig-subset-distribution)__).
+Of note, neutrophil expansion [@Silvin2020; @Penttila2021; @Vitte2020; @Laing2020; @Christensen2021] along with a profound reduction of non-classical monocyte levels [@Gatti2020; @Penttila2021; @Silvin2020; @Geanon2021; @Schulte-Schrepping2020; @Trombetta2021; @Mueller2021; @Zingaropoli2021; @Christensen2021; @Vasse2021; @Lucas2020] has been described as a characteristic of acute COVID-19.</span>
 
 The neutrophil CD64 was virtually absent in healthy controls (median $\Delta$MFI: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'neutrophils_CD64')`) but highly expressed in moderate (`r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'neutrophils_CD64')`) and severe COVID-19 (`r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'neutrophils_CD64')`).
-A similar pattern of CD64 regulation was discerned for classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD64')`) and intermediate monocytes (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD64')`), which, however, expressed this FC$\gamma$ receptor at high levels also in non-infected controls <span custom-style = 'revision'>(_Supplementary Figure S7__)</span>.
+A similar pattern of CD64 regulation was discerned for classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD64')`) and intermediate monocytes (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD64')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD64')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD64')`), which, however, expressed this FC$\gamma$ receptor at high levels also in non-infected controls <span custom-style = 'revision'>(__Supplementary Figure S7__)</span>.
 Surface expression of CD86 in classical (median $\Delta$MFI, healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'classicalMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'classicalMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'classicalMono_CD86')`) and intermediate monocytes (healthy: `r my_median(filter(mono_data$data_tbl, sev_paper == 'healthy'), 'intermediateMono_CD86')`, moderate: `r my_median(filter(mono_data$data_tbl, sev_paper == 'moderate'), 'intermediateMono_CD86')`, severe: `r my_median(filter(mono_data$data_tbl, sev_paper == 'severe'), 'intermediateMono_CD86')`) reached the maximum in moderate COVID-19 patients as compared with healthy controls and severe COVID-19 patients <span custom-style = 'revision'>(__Supplementary Figure S8__)</span>.
@@
 <span custom-style = 'revision'>
-Healthy controls and moderate COVID-19 cases comprised 75% of the largest participant cluster #1 and was characterized by the lowest levels CD40/CD64/CD71 parameters as well as healthy-like monocyte subset distribution, neutrophil levels and neutrophil CD64 expression (__Figure \@ref(fig:fig-clustering)AB__).
+Healthy controls and moderate COVID-19 cases comprised 75% of the largest participant cluster #1 that was characterized by the lowest levels CD40/CD64/CD71 parameters as well as healthy-like monocyte subset distribution, neutrophil levels and neutrophil CD64 expression (__Figure \@ref(fig:fig-clustering)AB__).
 Interestingly, none of the severe COVID-19 cases assigned to the cluster #1 required an ICU treatment (__Supplementary Table S4__).
-The remaining clusters consisted predominantly of severe COVID-19 patients.
-The hallmark of the cluster #2 was an increased expression of CD86, C163, the immunosuppresive PD-L1 molecule (CD279) and the iron exporter FPN1 in classical and intermediate monocytes as well as an expansion of the intermediate subset.
+The remaining clusters consisted predominantly of severe COVID-19 patients with increasing frequency.
+The hallmark of the cluster #2 was an increased expression of CD86, C163, the immunosuppressive PD-L1 molecule (CD279) and the iron exporter FPN1 in classical and intermediate monocytes as well as an expansion of the intermediate subset.
 The participant cluster #3 in turn was primarily characterized by an elevated neutrophil and monocyte CD64 as well as a depletion of the non-classical monocytes.
@@
 <span custom-style = 'revision'>
-In a comparison restricted to the COVID-19 individuals, the participant clusters demonstrated similar age and body mass index.
+In comparative analysis of COVID-19 individuals, the participant clusters demonstrated similar age and body mass index.
 In turn, the cluster #1 and #4 COVID-19 individuals were predominantly male (78%) as compared with the remaining subjects.
@@
 In agreement with previous <span custom-style = 'revision'>reports</span>, we observed elevated circulating IL6, CRP and neopterin levels and an expansion of neutrophils with a concomitant reduction in lymphocytes <span custom-style = 'revision'>in COVID-19 patients requiring respiratory support</span>.
-Similar to other infections [@Weiss2018; @Brandtner2020], this resulted in reduced levels of circulating iron and <span custom-style = 'revision'>elevated ferritin</span>, in line with few published reports <span custom-style = 'revision'>on COVID-19</span> [@Yagc2021; @Christensen2021; @Yagc2021; @Bellmann-Weiler2020; @Zhou2020].
+Similar to other infections [@Weiss2018; @Brandtner2020], this resulted in reduced levels of circulating iron and <span custom-style = 'revision'>elevated ferritin</span>, in line with few published reports <span custom-style = 'revision'>on COVID-19</span> [@Yagc2021; @Christensen2021; @Yagc2021; @Bellmann-Weiler2020; @Zhou2020; @Lanser2021].
 The sole cellular iron exporter FPN1 in monocyte and macrophages is down-regulated by the hormone hepcidin and diverse <span custom-style = 'revision'>inflammatory cytokines and up-regulated by cellular iron</span> [@Weiss2018; @Haschka2019; @Haschka2021; @Nemeth2004a].
-Interestingly, hypoferremia in acute COVID-19 was accompanied by only mildly elevated blood hepcidin [@Yagc2021; @Zhou2020]<span custom-style = 'revision'>, which may explain the consistently elevated FPN1 levels in the subsets of inflammatory COVID-19 patients assigned to the participant cluster #2.</span>
-<span custom-style = 'revision'>In general, the changes of the cellular iron trafficking and systemic turnover in COVID-19</span> may bear relevance for the effective SARS-CoV-2 control because iron levels significantly impact on differentiation of T cells, monocytes and macrophages and also affect anti-microbial immune effector mechanisms [@Weiss2018; @Haschka2021].
-<span custom-style = 'revision'>As such, these processes require further mechnistic and clinical investigations.</span>
+Interestingly, hypoferremia in acute COVID-19 was accompanied by only mildly elevated blood hepcidin levels [@Yagc2021; @Zhou2020]<span custom-style = 'revision'>, which may explain the consistently elevated FPN1 levels in the subsets of inflammatory COVID-19 patients assigned to the participant cluster #2.</span>
+<span custom-style = 'revision'>In general, the changes of the cellular iron trafficking and systemic turnover in COVID-19</span> may bear relevance for the effective SARS-CoV-2 control because iron levels significantly impact on differentiation of T cells and monocytes and <span custom-style = 'revision'>on macrophage polarization</span> and also affect anti-microbial immune effector mechanisms [@Weiss2018; @Haschka2021].
+<span custom-style = 'revision'>Furthermore, intracellular iron availability may promote viral replication [@Drakesmith2008].
+As such, the interplay of these processes requires further mechanistic and clinical investigations.</span>
 
@@
 
-<span custom-style = 'revision'>Our immunophenotyping results</span> revealed an up-regulation of the FC$\gamma$ receptor CD64 in neutrophils, classical and intermediate monocytes <span custom-style = 'revision'>in COVID-19 patients</span> as described recently [@Geanon2021; @Vitte2020; @Silvin2020; @Schulte-Schrepping2020; @Bourgoin2021].
-<span custom-style = 'revision'>For neutrophils, the elevated CD64 expression together with lowered neutrophil CD14 levels observed in the study group of severe COVID-19 patients may indicate the presence of less mature pro-neutrophil-like CD64^+^ CD14^low^ cells expanding in acute COVID-19 [@Schulte-Schrepping2020] and sepsis [@Meghraoui-Kheddar2020; @Sack2017] and hence represent a general reaction to a strong inflammatory stimuli triggering the emergency myelopoiesis [@Schulte-Schrepping2020; @Silvin2020].
+<span custom-style = 'revision'>Our immune-phenotyping results</span> revealed an up-regulation of the FC$\gamma$ receptor CD64 in neutrophils, classical and intermediate monocytes <span custom-style = 'revision'>in COVID-19 patients</span> as described recently [@Geanon2021; @Vitte2020; @Silvin2020; @Schulte-Schrepping2020; @Bourgoin2021].
+<span custom-style = 'revision'>For neutrophils, the elevated CD64 expression together with lowered neutrophil CD14 levels observed in the study group of severe COVID-19 patients may indicate the presence of less mature pro-neutrophil-like CD64^+^ CD14^low^ cells expanding in early COVID-19 [@Schulte-Schrepping2020] and sepsis [@Meghraoui-Kheddar2020; @Sack2017], and hence represent a general reaction to a strong inflammatory stimuli triggering emergency myelopoiesis [@Schulte-Schrepping2020; @Silvin2020].
 This is further supported by our clustering data, where the high neutrophil CD64 levels co-occurred with the neutrophil expansion in the highly inflammatory, severe COVID-19 dominated participant cluster #4.
@@
 <span custom-style = 'revision'>
-In the clustering analysis of our cytometry data, monocyte CD86 expression was found co-regulated with the immunosuppressive ligand CD279 (PD-L1), the alternative macrophage activation marker CD163 and the iron exporter FPN1 and was the highest in the participant cluster #2.
+In the clustering analysis of our cytometry data, monocyte CD86 expression was found to be co-regulated with the immunosuppressive ligand CD279 (PD-L1), the alternative macrophage activation marker CD163 and the iron exporter FPN1, the latter being highest in the participant cluster #2.
 Collectively, this suggests, that in a fraction of COVID-19 patients, subsumed under the participant cluster #2 in the study cohort, the blood monocytes may acquire an immunoregulatory, T cell-suppressive, iron export CD86^high^ CD279^high^ FPN1^high^ phenotype.
@@
 <span custom-style = 'revision'>
-To get a thorough overview of myeloid leukocyte response in acute COVID-19, we performed a clustering analysis with the self-organizing map technique [@Kohonen1995; @Vesanto2000] and could identify four distinct clusters of the study participants.
+To get a thorough overview of myeloid leukocyte response in acute COVID-19, we performed a clustering analysis with the self-organizing map technique [@Kohonen1995; @Vesanto2000] by which we could identify four distinct clusters among the study participants.
 The largest, healthy-like cluster #1 was associated with near normal composition and surface phenotype of neutrophils and monocytes as well as the lowest levels of systemic inflammation, largely normal iron homeostasis and low risk of ICU admission.
-the other clusters were characterized by a clearly inflammatory and iron-restrictive background but differed in the myeloid leukocyte phenotype.
+The other clusters were characterized by a clearly inflammatory and iron-restrictive phenotype but differed in the myeloid leukocyte characteristic.
 In particular, the cluster #2 was hallmarked by the presence of CD86-, CD163-, CD279- and FPN1-expressing monocytes suggestive of an immunoregulatory phenotype [@Silvin2020; @Penttila2021; @Christensen2021; @Schulte-Schrepping2020; @Bost2021] as discussed above.
@@
 Our study corroborates the massive changes in the neutrophil and monocyte compartment in acute moderate and severe COVID-19 including expansion of neutrophils, depletion of the non-classical monocytes
-<span custom-style = 'revision'>as well as regulation of CD64 and CD86 expression in myeloid leukocytes.
-Finally, we demonstrate the presence of four patterns of the innate myeloid response to SARS-CoV-2 with possible implication for diagnosis, risk assessment and therapy of COVID-19.</span>.
+<span custom-style = 'revision'>as well as regulation of CD64 and CD86 expression and altered iron trafficking in myeloid leukocytes.
+Finally, we demonstrate the presence of four patterns of the innate myeloid response to SARS-CoV-2 with possible implication for diagnosis, risk assessment and therapy of COVID-19.</span>
 
@@
 
-he study was performed between 21.03.2020 and 16.11.2020 at the hospital of the University of Innsbruck, Tyrol, Austria.
+The study was performed between 21.03.2020 and 16.11.2020 at the hospital of the University of Innsbruck, Tyrol, Austria.
 The inclusion criteria for COVID-19 participants were age $\geq$ 18 years and hospital admission because of acute SARS-CoV-2 infection confirmed by nasal or oral swab PCR test.
@@
 Whole venous EDTA blood (5 - 10 ml) for the flow cytometry measurement was withdrawn within 72 hours from hospital admission.
+
+## Measurement of laboratory inflammation and iron markers
+
+<span custom-style = 'revision'>CRP in plasma was determined by particle-enhanced immunological turbidity assay: human CRP agglutinated with latex particles coated with anti-CRP monoclonal antibody (mouse).
+Aggregates were determined turbidimetrically.
+IL-6 was assessed in plasma by electro-chemiluminescence assay (ECLIA) in sandwich principle.
+Neopterin was analyzed by a competitive enzyme immunoassay for the direct quantitative determination of neopterin.
+Quantitative determination of ferritin in human plasma was performed using latex particle-enhanced immunological turbidity assay.
+The amount of iron in human plasma was measured photometrically by FerroZine method without deproteinization.
+Transferrin saturation was measured by an immunological turbidity test.</span>
 
@@
 The cytometry data was analyzed by an experienced operator blinded to the study group assignment with a strategy depicted in __Supplementary Figures S1__ and __S2__. 
-<span custom-style = 'revision'>Neutrophils were identified within the lineage-negative leukocytes (CD3^-^ CD19^-^ CD56^-^) by logical AND gating of CD11b^+^, CD16^+^, CD62L^+^ and SSC^lo^ events.
+<span custom-style = 'revision'>Neutrophils were identified within the lineage-negative leukocytes (CD3^-^ CD19^-^ CD56^-^) by logical gating (AND) of CD11b^+^, CD16^+^, CD62L^+^ and SSC^lo^ events.
 Monocytes were defined within the non-neutrophil gate (logical NOT gating) by UMAP [@McInnes2018] in respect to CD11b, CD14, C15, CD16, CD62L, CCR2, CX3CR1 and HLA-DR expression.
@@
 Monocyte: lymphocyte ratio was calculated as a ratio of the CD45^+^ percents of HLA-DR^+^ pan-monocytes to Lineage^+^ (CD3, CD19, CD56) cells.
-Neutrophil: lymphocyte ratio was calculated as a ratio of the CD45^+^ percents of CD15^+^ SSC^hi^ neutrophils to Lineage^+^ cells.
+Neutrophil: lymphocyte ratio was calculated as a ratio of the CD45^+^ percent of CD15^+^ SSC^hi^ neutrophils to Lineage^+^ cells.
 
@@
 
-Statistical analysis was done with R version 4.0.5 and packages, _tidyverse_, <span custom-style = 'revision'>_DescTools_, _rstatix_</span> and _cowplot_ packages [@Wickham2019; @Wilke2019], <span custom-style = 'revision'>the in-house-developed package _ExDA_ (https://github.com/PiotrTymoszuk/ExDA) and a script with clustering function wrappers (https://github.com/PiotrTymoszuk/clustering-tools-2).</span>
+Statistical analysis was done with R version 4.0.5 and packages, _tidyverse_, <span custom-style = 'revision'>_DescTools_, _rstatix_</span> and _cowplot_ packages [@Wickham2019; @Wilke2019], <span custom-style = 'revision'>the in-house-developed packages _ExDA_ (explorative data analysis and hypothesis testing, https://github.com/PiotrTymoszuk/ExDA) and _clustTools_ (wrappers for clustering and self-organizing maps, https://github.com/PiotrTymoszuk/clustTools).</span>
 
@@
 
-PT owns his data science enterprise, Data Analytics as a Service Tirol.
-PT has also received an honorarium for the study data management, curation and analysis.
-The author declares no other conflicts of interest.
-Other authors declare that non conflict of interest exists.
+PT now owns his data science enterprise, Data Analytics as a Service Tirol. The author declares no other conflicts of interest. Other authors declare that non conflict of interest exists.
 

### 2022-03-25 analysis scripts/cluster_characteristic.R

@@ -29,6 +29,7 @@
   ## n numbers
   
   clust_chara$n_numbers <- clust_chara$analysis_tbl %>% 
+    filter(sev_paper %in% c('moderate', 'severe')) %>% 
     count(clust_id)
   
   
@@ -80,6 +81,7 @@
                  clust_4 = clust_chara$n_numbers$n[4]), .) %>% 
     map_dfc(stri_replace, regex = '\\nComplete.*$', replacement = '') %>% 
     map_dfc(stri_replace, regex = 'no:.*\\nyes:\\s{1}', replacement = '') %>% 
+    map_dfc(stri_replace, regex = '^no: 100%.*', replacement = '0% (0)') %>% 
     map_dfc(stri_replace_all, fixed = '% (', replacement = '% (n = ')
   
 # Kruskal-Wallis and Chi-squared tests ------
@@ -130,7 +132,7 @@
                                  clust_chara$result_tbl[c('variable', 'significance_post')]) %>% 
     reduce(left_join, by = 'variable') %>% 
     mutate(variable = translate_var(variable, type = 'plot_lab'), 
-           variable = ifelse(is.na(variable), 'COVID-19 severity', variable))
+           variable = ifelse(is.na(variable), 'N COVID-19 patients', variable))
   
 # Summary plots: inflammatory and iron parameters ------
   

### 2022-03-25 analysis scripts/clustering.R

@@ -62,7 +62,7 @@
   
   cyto_clust$ft_diagnostic <- list(diagn_plots = function(x) plot(x, cust_theme = proj_globals$common_theme), 
                                    variance = var, 
-                                   cv = function(x) cv(x, nfolds = 10, nearest_n = 5)) %>% 
+                                   cv = function(x) cv(x, nfolds = 10, kNN = 5)) %>% 
     map(~.x(cyto_clust$ft_object))
   
   ## renaming of the clusters: reader-friendly names
@@ -98,8 +98,8 @@
   ## variance and cross-validation error
   
   cyto_clust$part_diagnostic <- list(diagn_plots = function(x) plot(x, cust_theme = proj_globals$common_theme), 
-                                   variance = var, 
-                                   cv = function(x) cv(x, nfolds = 10, nearest_n = 5)) %>% 
+                                     variance = var, 
+                                     cv = function(x) cv(x, nfolds = 10, kNN = 5)) %>% 
     map(~.x(cyto_clust$part_object))
   
   ## renaming of the clusters
@@ -119,8 +119,8 @@
   
   insert_msg('Plotting of the clustering results as a heat map')
   
-  cyto_clust$heat_map <- plot_clust_hm(sample_clust_object = cyto_clust$part_object, 
-                                       feature_clust_object = cyto_clust$ft_object, 
+  cyto_clust$heat_map <- plot_clust_hm(x_object = cyto_clust$part_object, 
+                                       y_object = cyto_clust$ft_object, 
                                        cust_theme = proj_globals$common_theme, 
                                        plot_title = 'Participant and cytometry feature clustering', 
                                        plot_subtitle = 'Self-organizing map/HCl', 

### 2022-03-25 paper_analyses.R

@@ -5,10 +5,11 @@
   insert_head()
 
   library(exda)
+  library(somKernels)
+  library(clustTools)
 
   c('globals_setup.R', 
-    './tools/project_tools.R', 
-    './tools/clust_tools2.R') %>% 
+    './tools/project_tools.R') %>% 
     source_all(message = TRUE, crash = TRUE)
 
 # executable scripts ------
